---
title: "Algo5 noise variants: identifiability and convergence"
format: html
execute:
  cache: true
---

```{r cache = FALSE}
#| label: init
#| message: false
#| cache: false
library(bmm)
library(ggplot2)
library(tidyr)
library(dplyr)

source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: {.callout-note}
## Prerequisites
This notebook follows up on the [stochastic algorithms section](reinforcement-learning-arc.html#stochastic-algorithms) of the RL exploration notebook. The identifiability problem diagnosed here is documented in `meta/2026-02-10_algo5-identifiability-analysis.md`.
:::

# Motivation

The noisy-gradient model (algo5) with gradient-scaled noise has good parameter recovery in simulation but shows **bimodal posterior behavior** when fitted to real data via MCMC. Chains get stuck exploring one of two modes:

| Mode | $\tau$ | $\delta_0$ | $\sigma_g$ |
|------|--------|-----------|-----------|
| A    | high   | high      | low       |
| B    | low    | low       | high      |

This arises from a **three-way multiplicative ridge** in the likelihood. The arc transition is:

$$
x_{t+1} \sim \mathcal{N}\!\left(x_t + \delta_t\, g_t(\tau),\; (\delta_t\,\sigma_g)^2\right)
$$

where $g_t \approx C/\tau$ near the decision boundary, so the mean depends on $\delta_0/\tau$ and the variance on $(\delta_0 \sigma_g)^2$. The approximate invariance

$$
(\tau, \delta_0, \sigma_g) \to (\lambda\tau,\; \lambda\delta_0,\; \sigma_g/\lambda)
$$

preserves both, creating a ridge that MCMC struggles to traverse.

This notebook investigates **alternative noise formulations** that maintain the gradient-ascent spirit of algo5 while improving estimation geometry. For each variant we check convergence behavior in forward simulation --- a prerequisite before attempting estimation.

# Reference: current model (gradient-scaled noise)

The current model scales noise by the learning rate: $\text{SD} = \delta_t \sigma_g$. Because the noise decays with $\delta_t$, even large $\sigma_g$ does not prevent convergence.

```{r}
#| label: ref-noisy-grad-ribbon
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    deterministic = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0),
    sigma_g_0.5   = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0.5),
    sigma_g_1     = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 1),
    sigma_g_2     = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2),
    sigma_g_5     = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 5)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: ref-noisy-grad-spaghetti
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(noisy_grad = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2)),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

# Proposal 1: Constant noise

Replace $\delta_t \sigma_g$ with a constant $\sigma_w$:

$$
x_{t+1} = x_t + \delta_t\, g_t(\tau) + \sigma_w\, \varepsilon_t
$$

This completely decouples $\sigma_w$ from $\delta_0$, breaking the variance ridge. However, since $\sigma_w$ does not decay, the noise dominates late-trial dynamics where $\delta_t \to 0$. This is equivalent to the "logit noise" variant explored earlier, which was shown to destroy algo5's convergence.

```{r}
#| label: p1-const-ribbon
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    deterministic  = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0),
    sigma_w_0.01   = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.01),
    sigma_w_0.05   = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.05),
    sigma_w_0.1    = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.1),
    sigma_w_0.2    = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.2)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: p1-const-spaghetti
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(const_noise = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.1)),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

As expected, constant noise prevents convergence to a stable arc. Late-trial variability remains high because the noise floor $\sigma_w$ overwhelms the vanishing gradient step. This confirms the earlier finding from the logit-noise experiments. The model would be identifiable for estimation but generates unrealistic trajectories.

# Proposal 2: Langevin noise scaling

Use $\sqrt{\delta_t}$ scaling instead of $\delta_t$:

$$
x_{t+1} = x_t + \delta_t\, g_t(\tau) + \sqrt{\delta_t}\;\sigma_L\, \varepsilon_t
$$

The noise still decays (preserving convergence) but more slowly than the gradient step. Crucially, $\delta_0$ enters the mean **linearly** and the variance as $\delta_0^{1/2}$, so the variance ridge $\delta_0 \sigma_g = \text{const}$ is replaced by $\delta_0^{1/2} \sigma_L = \text{const}$ --- a much tighter constraint that breaks the three-way invariance.

```{r}
#| label: p2-langevin-ribbon
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    deterministic  = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0),
    sigma_L_0.5    = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 0.5),
    sigma_L_1      = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 1),
    sigma_L_2      = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 2),
    sigma_L_5      = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 5)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: p2-langevin-spaghetti
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(langevin = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 2)),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

## Direct comparison: noise scaling regimes

Compare all three noise scalings at matched effective noise levels. For a fair comparison at a "typical" trial ($t = 100$, $\delta_0 = 0.1$): $\delta_t \approx 0.01$, so we match:

- Gradient-scaled: $\delta_t \sigma_g = 0.01 \times 2 = 0.02$
- Langevin: $\sqrt{\delta_t}\, \sigma_L = 0.1 \times 0.2 = 0.02$
- Constant: $\sigma_w = 0.02$

```{r}
#| label: p2-comparison-ribbon
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    `grad-scaled (sig_g=2)` = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2),
    `langevin (sig_L=0.2)`  = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 0.2),
    `constant (sig_w=0.02)` = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.02)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

# Proposal 3: Reparametrize the existing model

This proposal keeps the original generative model **exactly** but estimates $(\phi, \omega, \tau)$ instead of $(\tau, \delta_0, \sigma_g)$:

$$
\phi = \delta_0/\tau, \quad \omega = \delta_0 \cdot \sigma_g
$$

Since the generative model is identical, simulations are the same as the reference. What changes is the posterior geometry during estimation. The ridge direction $(\lambda\tau, \lambda\delta_0, \sigma_g/\lambda)$ maps to $(\lambda\tau, \phi, \omega)$ --- only $\tau$ moves while $\phi$ and $\omega$ stay fixed. This concentrates the degeneracy in one dimension.

No simulation needed here --- we plot the original model's ridge structure instead.

```{r}
#| label: p3-ridge-visualization
#| fig-width: 10
#| fig-height: 4

lambda_grid <- seq(0.2, 5, length.out = 200)

tau0 <- 0.025; delta0_0 <- 0.1; sigma_g0 <- 2

original <- tibble(
  lambda = lambda_grid,
  tau    = tau0 * lambda_grid,
  delta0 = delta0_0 * lambda_grid,
  sigma_g = sigma_g0 / lambda_grid,
  phi    = delta0_0 / tau0,
  omega  = delta0_0 * sigma_g0
)

p1 <- ggplot(original, aes(x = tau, y = delta0)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(original, lambda == 1), color = "red", size = 3) +
  labs(title = "Original: tau-delta ridge", x = expression(tau), y = expression(delta[0]))

p2 <- ggplot(original, aes(x = delta0, y = sigma_g)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(original, lambda == 1), color = "red", size = 3) +
  labs(title = "Original: delta-sigma_g ridge", x = expression(delta[0]), y = expression(sigma[g]))

p3 <- ggplot(original, aes(x = tau, y = phi)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(original, lambda == 1), color = "red", size = 3) +
  labs(title = "Reparametrized: tau vs phi", x = expression(tau), y = expression(phi == delta[0]/tau))

p4 <- ggplot(original, aes(x = tau, y = omega)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(original, lambda == 1), color = "red", size = 3) +
  labs(title = "Reparametrized: tau vs omega", x = expression(tau), y = expression(omega == delta[0] %.% sigma[g]))

library(patchwork)
(p1 | p2) / (p3 | p4)
```

In the original parametrization, the ridge sweeps through both $\tau$--$\delta_0$ and $\delta_0$--$\sigma_g$ panels (top row). In the reparametrized space (bottom row), $\phi$ and $\omega$ are **flat** along the ridge --- MCMC only needs to marginalize over $\tau$.

# Proposal 4: Fix $\tau$

The temperature $\tau$ controls gradient smoothness but has little direct scientific meaning. Fixing it at a small value eliminates one parameter and the $\tau$--$\delta_0$ coupling entirely. The model reduces to a two-parameter arc transition $(c, \kappa, \delta_0, \sigma_g)$ with the same geometry as algo4.

We verify that convergence behavior is insensitive to the choice of $\tau$:

```{r}
#| label: p4-fix-tau-ribbon
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    `tau=0.005` = algo5_noisy_grad(tau = 0.005, delta0 = 0.1, sigma_g = 2),
    `tau=0.01`  = algo5_noisy_grad(tau = 0.01,  delta0 = 0.1, sigma_g = 2),
    `tau=0.025` = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2),
    `tau=0.05`  = algo5_noisy_grad(tau = 0.05,  delta0 = 0.1, sigma_g = 2),
    `tau=0.1`   = algo5_noisy_grad(tau = 0.1,   delta0 = 0.1, sigma_g = 2)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: p4-fix-tau-spaghetti
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(
    `tau=0.01`  = algo5_noisy_grad(tau = 0.01, delta0 = 0.1, sigma_g = 2),
    `tau=0.025` = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2)
  ),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

If convergence and asymptotic behavior are similar across a range of $\tau$ values, the parameter adds little information and can safely be fixed.

# Proposal 5: Simpler stochastic arc model

Instead of computing the reward gradient via a smooth approximation, use an **error-correction** rule on the logit scale. The arc tracks toward the observed error magnitude:

$$
x_{t+1} = x_t + \delta_t\,\bigl[\operatorname{logit}(|y_t|/\pi) - x_t\bigr] + \sigma_w\,\varepsilon_t
$$

This is an exponential smoother (leaky integrator) on the logit scale. Parameters: $(\delta_0, \sigma_w)$ --- no $\tau$, no gradient. The agent adjusts its arc toward wherever the last error landed in logit space.

The tracked target $E[\operatorname{logit}(|y|/\pi)]$ is not the reward-optimal arc, but the two are correlated through the SDM error distribution.

```{r}
#| label: p5-error-correction-ribbon
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    deterministic  = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0),
    `ec_sw=0`      = algo_error_correction(delta0 = 0.1, sigma_w = 0),
    `ec_sw=0.05`   = algo_error_correction(delta0 = 0.1, sigma_w = 0.05),
    `ec_sw=0.1`    = algo_error_correction(delta0 = 0.1, sigma_w = 0.1),
    `ec_sw=0.2`    = algo_error_correction(delta0 = 0.1, sigma_w = 0.2)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: p5-error-correction-spaghetti
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(error_correction = algo_error_correction(delta0 = 0.1, sigma_w = 0.1)),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

## Compare error-correction with algo4

The error-correction model is conceptually similar to algo4 (both track $|y_t|$ rather than computing a gradient), but algo4 operates on the native scale while error-correction operates on the logit scale. Compare them head-to-head:

```{r}
#| label: p5-vs-algo4-ribbon
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    algo4         = algo4(delta = 0.1),
    `algo5 (det)` = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0),
    `ec (det)`    = algo_error_correction(delta0 = 0.1, sigma_w = 0)
  ),
  cs = c(2, 4, 8, 16),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

# Comparison across all proposals: Langevin vs noise-gradient vs fixed-tau

The proposals that maintain convergence (Langevin, reparametrize, fix-tau) produce identical or nearly identical forward trajectories --- they differ only in estimation geometry. Compare the two models with genuinely different dynamics:

```{r}
#| label: all-proposals-ribbon
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    `noisy-grad`    = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2),
    `langevin`      = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 2),
    `const-noise`   = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.05),
    `error-correct` = algo_error_correction(delta0 = 0.1, sigma_w = 0.1)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

# Summary

| Variant | Noise SD | Converges? | Ridge broken? | Parameters |
|---------|----------|-----------|--------------|------------|
| Current (gradient-scaled) | $\delta_t \sigma_g$ | Yes | No — three-way ridge | $\tau, \delta_0, \sigma_g$ |
| Proposal 1 (constant) | $\sigma_w$ | No | Yes — fully | $\tau, \delta_0, \sigma_w$ |
| Proposal 2 (Langevin) | $\sqrt{\delta_t}\,\sigma_L$ | Yes | Partially — weakened | $\tau, \delta_0, \sigma_L$ |
| Proposal 3 (reparametrize) | $\delta_t \sigma_g$ | Yes | Concentrated in $\tau$ | $\phi, \omega, \tau$ |
| Proposal 4 (fix $\tau$) | $\delta_t \sigma_g$ | Yes | Yes — $\tau$ eliminated | $\delta_0, \sigma_g$ |
| Proposal 5 (error-correction) | $\sigma_w$ | Depends on $\sigma_w$ | N/A — no $\tau$ | $\delta_0, \sigma_w$ |

**Recommendations:**

1. **Proposal 2 (Langevin)** is the best single change --- it preserves convergence and substantially weakens the ridge.
2. **Proposal 4 (fix $\tau$)** is the simplest fix if $\tau$ sensitivity analysis confirms stability.
3. **Proposal 3 (reparametrize)** can be combined with any noise variant to further improve estimation.
4. **Proposal 5 (error-correction)** is worth exploring as a simpler alternative that avoids the gradient machinery entirely.
