---
title: "Algo5 noise variants: identifiability and convergence"
format: html
execute:
  cache: true
editor:
  render-on-save: true
---

```{r}
#| label: init
#| message: false
#| cache: false
library(bmm)
library(ggplot2)
library(tidyr)
library(dplyr)

source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: {.callout-note}
## Prerequisites
This notebook follows up on the [stochastic algorithms section](reinforcement-learning-arc.html#stochastic-algorithms) of the RL exploration notebook. The identifiability problem diagnosed here is documented in `meta/2026-02-10_algo5-identifiability-analysis.md`.
:::

# Motivation

The noisy-gradient model (algo5) with gradient-scaled noise has good parameter recovery in simulation but shows **bimodal posterior behavior** when fitted to real data via MCMC. Chains get stuck exploring one of two modes:

| Mode | $\tau$ | $\delta_0$ | $\sigma_g$ |
|------|--------|-----------|-----------|
| A    | high   | high      | low       |
| B    | low    | low       | high      |

This arises from a **three-way multiplicative ridge** in the likelihood. The arc transition is:

$$
x_{t+1} \sim \mathcal{N}\!\left(x_t + \delta_t\, g_t(\tau),\; (\delta_t\,\sigma_g)^2\right)
$$

where $g_t \approx C/\tau$ near the decision boundary, so the mean depends on $\delta_0/\tau$ and the variance on $(\delta_0 \sigma_g)^2$. The approximate invariance

$$
(\tau, \delta_0, \sigma_g) \to (\lambda\tau,\; \lambda\delta_0,\; \sigma_g/\lambda)
$$

preserves both, creating a ridge that MCMC struggles to traverse.

This notebook investigates **alternative noise formulations** that maintain the gradient-ascent spirit of algo5 while improving estimation geometry. For each variant we check convergence behavior in forward simulation --- a prerequisite before attempting estimation.

# Reference: current model (gradient-scaled noise)

The current model scales noise by the learning rate: $\text{SD} = \delta_t \sigma_g$. Because the noise decays with $\delta_t$, even large $\sigma_g$ does not prevent convergence.

```{r}
#| label: ref-noisy-grad-ribbon
#| fig-asp: 0.8
#| fig-width: 10
ribbon_summary_runs(
  alg_list = list(
    deterministic = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0),
    sigma_g_0.5   = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0.5),
    sigma_g_1     = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 1),
    sigma_g_2     = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2),
    sigma_g_5     = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 5)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: ref-noisy-grad-spaghetti
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(noisy_grad = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2)),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

# Proposal 1: Constant noise

Replace $\delta_t \sigma_g$ with a constant $\sigma_w$:

$$
x_{t+1} = x_t + \delta_t\, g_t(\tau) + \sigma_w\, \varepsilon_t
$$

This completely decouples $\sigma_w$ from $\delta_0$, breaking the variance ridge. However, since $\sigma_w$ does not decay, the noise dominates late-trial dynamics where $\delta_t \to 0$. This is equivalent to the "logit noise" variant explored earlier, which was shown to destroy algo5's convergence.

```{r}
#| label: p1-const-ribbon
#| fig-asp: 0.8
#| fig-width: 10
ribbon_summary_runs(
  alg_list = list(
    deterministic  = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0),
    sigma_w_0.01   = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.01),
    sigma_w_0.05   = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.05),
    sigma_w_0.1    = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.1),
    sigma_w_0.2    = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.2)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: p1-const-spaghetti
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(const_noise = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.1)),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

As expected, constant noise prevents convergence to a stable arc. Late-trial variability remains high because the noise floor $\sigma_w$ overwhelms the vanishing gradient step. This confirms the earlier finding from the logit-noise experiments. The model would be identifiable for estimation but generates unrealistic trajectories.

# Proposal 2: Langevin noise scaling

Use $\sqrt{\delta_t}$ scaling instead of $\delta_t$:

$$
x_{t+1} = x_t + \delta_t\, g_t(\tau) + \sqrt{\delta_t}\;\sigma_L\, \varepsilon_t
$$

The noise still decays but more slowly than the gradient step. Crucially, $\delta_0$ enters the mean **linearly** and the variance as $\delta_0^{1/2}$, so the variance ridge $\delta_0 \sigma_g = \text{const}$ is replaced by $\delta_0^{1/2} \sigma_L = \text{const}$ --- a much tighter constraint that breaks the three-way invariance. But sigma will have to be much smaller, otherwise convergence is once again destroyed.

```{r}
#| label: p2-langevin-ribbon
#| fig-asp: 0.8
#| fig-width: 10
ribbon_summary_runs(
  alg_list = list(
    deterministic  = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0),
    sigma_L_0.5    = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 0.5),
    sigma_L_1      = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 1),
    sigma_L_2      = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 2),
    sigma_L_5      = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 5)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

```{r}
#| label: p2-langevin-spaghetti
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(langevin = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 2)),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

## Direct comparison: noise scaling regimes

Compare all three noise scalings at matched effective noise levels. For a fair comparison at a "typical" trial ($t = 100$, $\delta_0 = 0.1$): $\delta_t \approx 0.01$, so we match:

- Gradient-scaled: $\delta_t \sigma_g = 0.01 \times 2 = 0.02$
- Langevin: $\sqrt{\delta_t}\, \sigma_L = 0.1 \times 0.2 = 0.02$
- Constant: $\sigma_w = 0.02$

Interestingly, there is little difference between them, as the plot below shows. Perhaps constant noise might still work as long as we keep it low.

```{r}
#| label: p2-comparison-ribbon
#| fig-asp: 0.8
#| fig-width: 10
ribbon_summary_runs(
  alg_list = list(
    `grad-scaled (sig_g=2)` = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2),
    `langevin (sig_L=0.2)`  = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 0.2),
    `constant (sig_w=0.02)` = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.02)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```


Here is how the resulting arc trajectory variance changes under the three models:

```{r}
#| label: p2-arc-distribution-variance-comparison
sigma_grid <- list(
  `grad-scaled` = c(0.5, 1, 2, 5),
  langevin      = c(0, 0.1, 0.2, 0.3, 0.4, 0.5),
  constant      = c(0, 0.01, 0.02, 0.03, 0.04, 0.05)
)

make_alg <- list(
  `grad-scaled` = \(s) algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = s),
  langevin      = \(s) algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = s),
  constant      = \(s) algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = s)
)

n_runs   <- 200
n_trials <- 1000
cs       <- c(4, 8)
kappa    <- 3

set.seed(42)

var_df <- purrr::map_dfr(names(sigma_grid), \(model) {
  purrr::map_dfr(sigma_grid[[model]], \(sig) {
    purrr::map_dfr(cs, \(c_val) {
      within_vars <- vapply(seq_len(n_runs), \(r) {
        a <- run_arc_algorithm(
          arc_update_fun = make_alg[[model]](sig),
          c = c_val, kappa = kappa,
          init_arc = pi / 2, n_trials = n_trials
        )$a
        var(a)
      }, numeric(1))
      tibble(
        model = model,
        sigma = sig,
        c     = c_val,
        var_a = within_vars
      )
    })
  })
})

var_df$sigma_label <- factor(var_df$sigma)
var_df$c_label     <- paste0("c = ", var_df$c)
```

```{r}
#| label: p2-arc-distribution-variance-comparison-plot
#| dependson: p2-arc-distribution-variance-comparison
#| fig-width: 12
#| fig-height: 10
ggplot(var_df, aes(x = sigma_label, y = var_a)) +
  geom_boxplot(outlier.size = 0.8) +
  facet_wrap(c_label ~ model, scales = "free") +
  labs(
    x = expression(sigma), y = "Within-run Var(arc width)",
    title = "Within-run arc variance by noise model and sigma"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

# Proposal 3: Reparametrize the existing model

This proposal keeps the original generative model **exactly** but estimates $(\phi, \omega, \tau)$ instead of $(\tau, \delta_0, \sigma_g)$:

$$
\phi = \delta_0/\tau, \quad \omega = \delta_0 \cdot \sigma_g
$$

Since the generative model is identical, simulations are the same as the reference. What changes is the posterior geometry during estimation. The ridge direction $(\lambda\tau, \lambda\delta_0, \sigma_g/\lambda)$ maps to $(\lambda\tau, \phi, \omega)$ --- only $\tau$ moves while $\phi$ and $\omega$ stay fixed. This concentrates the degeneracy in one dimension.

No simulation needed here --- we plot the original model's ridge structure instead.

```{r}
#| label: p3-ridge-visualization
#| fig-width: 10
#| fig-height: 4

lambda_grid <- seq(0.2, 5, length.out = 200)

tau0 <- 0.025; delta0_0 <- 0.1; sigma_g0 <- 2

original <- tibble(
  lambda = lambda_grid,
  tau    = tau0 * lambda_grid,
  delta0 = delta0_0 * lambda_grid,
  sigma_g = sigma_g0 / lambda_grid,
  phi    = delta0_0 / tau0,
  omega  = delta0_0 * sigma_g0
)

p1 <- ggplot(original, aes(x = tau, y = delta0)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(original, lambda == 1), color = "red", size = 3) +
  labs(title = "Original: tau-delta ridge", x = expression(tau), y = expression(delta[0]))

p2 <- ggplot(original, aes(x = delta0, y = sigma_g)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(original, lambda == 1), color = "red", size = 3) +
  labs(title = "Original: delta-sigma_g ridge", x = expression(delta[0]), y = expression(sigma[g]))

p3 <- ggplot(original, aes(x = tau, y = phi)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(original, lambda == 1), color = "red", size = 3) +
  labs(title = "Reparametrized: tau vs phi", x = expression(tau), y = expression(phi == delta[0]/tau))

p4 <- ggplot(original, aes(x = tau, y = omega)) +
  geom_line(linewidth = 1) +
  geom_point(data = filter(original, lambda == 1), color = "red", size = 3) +
  labs(title = "Reparametrized: tau vs omega", x = expression(tau), y = expression(omega == delta[0] %.% sigma[g]))

library(patchwork)
(p1 | p2) / (p3 | p4)
```

In the original parametrization, the ridge sweeps through both $\tau$--$\delta_0$ and $\delta_0$--$\sigma_g$ panels (top row). In the reparametrized space (bottom row), $\phi$ and $\omega$ are **flat** along the ridge --- MCMC only needs to marginalize over $\tau$.

# Proposal 4: Fix $\tau$

The temperature $\tau$ controls gradient smoothness but has little direct scientific meaning. Fixing it at a small value eliminates one parameter and the $\tau$--$\delta_0$ coupling entirely. The model reduces to a two-parameter arc transition $(c, \kappa, \delta_0, \sigma_g)$ with the same geometry as algo4.

We verify that convergence behavior is mostly insensitive to the choice of $\tau <= 0.05$:

```{r}
#| label: p4-fix-tau-ribbon
#| fig-asp: 0.8
#| fig-width: 10
ribbon_summary_runs(
  alg_list = list(
    `tau=0.005` = algo5_noisy_grad(tau = 0.005, delta0 = 0.1, sigma_g = 2),
    `tau=0.01`  = algo5_noisy_grad(tau = 0.01,  delta0 = 0.1, sigma_g = 2),
    `tau=0.025` = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2),
    `tau=0.05`  = algo5_noisy_grad(tau = 0.05,  delta0 = 0.1, sigma_g = 2),
    `tau=0.1`   = algo5_noisy_grad(tau = 0.1,   delta0 = 0.1, sigma_g = 2)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

If convergence and asymptotic behavior are similar across a range of $\tau$ values, the parameter adds little information and can safely be fixed.

The question is whether and how much $\tau$ adds to the observed variance of the arcs. Let's test that:


Here is how the resulting arc trajectory variance changes with tau:

```{r}
#| label: p4-arc-variance-tau-dependence
tau_grid <- seq(0.01, 0.05, 0.005)
n_runs   <- 200
n_trials <- 1000
cs       <- c(4, 8)
kappa    <- 3

set.seed(42)

var_df <- purrr::map_dfr(tau_grid, \(tau) {
  purrr::map_dfr(cs, \(c_val) {
    within_vars <- vapply(seq_len(n_runs), \(r) {
      a <- run_arc_algorithm(
        arc_update_fun = algo5_noisy_grad(tau = tau, delta0 = 0.1, sigma_g = 2),
        c = c_val, kappa = kappa,
        init_arc = pi / 2, n_trials = n_trials
      )$a
      var(a)
    }, numeric(1))
    tibble(tau = tau, c = c_val, var_a = within_vars)
  })
})

var_df$tau_label <- factor(var_df$tau)
var_df$c_label   <- paste0("c = ", var_df$c)
```

```{r}
#| label: p4-arc-variance-tau-dependence-plot
#| dependson: p4-arc-variance-tau-dependence
#| fig-width: 10
#| fig-height: 5
ggplot(var_df, aes(x = tau_label, y = var_a)) +
  geom_boxplot(outlier.size = 0.8) +
  facet_wrap(~c_label, scales = "free_y") +
  labs(
    x = expression(tau), y = "Within-run Var(arc width)",
    title = "Within-run arc variance by tau"
  ) +
  theme_minimal(base_size = 13)
```

# Comparison across all proposals: Langevin vs noise-gradient vs fixed-tau

The proposals that maintain convergence (Langevin, reparametrize, fix-tau) produce identical or nearly identical forward trajectories --- they differ only in estimation geometry. Compare the two models with genuinely different dynamics:

```{r}
#| label: all-proposals-ribbon
#| fig-asp: 0.8
#| fig-width: 10
ribbon_summary_runs(
  alg_list = list(
    `noisy-grad`    = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2),
    `langevin`      = algo5_langevin(tau = 0.025, delta0 = 0.1, sigma_L = 0.5),
    `const-noise`   = algo5_const_noise(tau = 0.025, delta0 = 0.1, sigma_w = 0.05)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

# Summary

| Variant | Noise SD | Converges? | Ridge broken? | Parameters |
|---------|----------|-----------|--------------|------------|
| Current (gradient-scaled) | $\delta_t \sigma_g$ | Yes | No — three-way ridge | $\tau, \delta_0, \sigma_g$ |
| Proposal 1 (constant) | $\sigma_w$ | No | Yes — fully | $\tau, \delta_0, \sigma_w$ |
| Proposal 2 (Langevin) | $\sqrt{\delta_t}\,\sigma_L$ | Yes | Partially — weakened | $\tau, \delta_0, \sigma_L$ |
| Proposal 3 (reparametrize) | $\delta_t \sigma_g$ | Yes | Concentrated in $\tau$ | $\phi, \omega, \tau$ |
| Proposal 4 (fix $\tau$) | $\delta_t \sigma_g$ | Yes | Yes — $\tau$ eliminated | $\delta_0, \sigma_g$ |
| Proposal 5 (error-correction) | $\sigma_w$ | Depends on $\sigma_w$ | N/A — no $\tau$ | $\delta_0, \sigma_w$ |

**Recommendations:**

1. **Proposal 2 (Langevin)** is the best single change --- it preserves convergence and substantially weakens the ridge.
2. **Proposal 4 (fix $\tau$)** is the simplest fix if $\tau$ sensitivity analysis confirms stability.
3. **Proposal 3 (reparametrize)** can be combined with any noise variant to further improve estimation.
4. **Proposal 5 (error-correction)** is worth exploring as a simpler alternative that avoids the gradient machinery entirely.
