---
title: "Counterfactual Q-Learning for Arc Width"
format: html
execute:
  cache: false
---

```{r cache = FALSE}
#| label: init
#| message: false
#| cache: false
library(bmm)
library(ggplot2)
library(tidyr)
library(dplyr)

source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: {.callout-note}
## Prerequisites
This notebook builds on the reinforcement learning exploration in [RL Algorithms](reinforcement-learning-arc.qmd). It introduces a Counterfactual Q-Learning (CRL) agent that learns a full value function over the arc-width action space, rather than applying local update heuristics.
:::

# Model Overview

Previous algorithms (algo4, algo5) use local update rules — they adjust the current arc based only on whether the last trial was a hit or miss (and by how much). The **Counterfactual Reinforcement Learning (CRL)** agent takes a fundamentally different approach: it maintains a value estimate $Q(\alpha)$ for *every* possible arc width, and updates *all* of them on every trial.

## Why "counterfactual"?

After observing the response error $y_t$, the agent knows the true target location. This means it can compute the reward it *would have received* for any arc width $\alpha$, not just the one it chose. This is the counterfactual: "what if I had set my arc to $\alpha$?"

## Discretization

The continuous action space $[0, \pi]$ is discretized into $N$ equally spaced values $(0, \pi/N, 2\pi/N, \dots, z_k, \dots, \pi)$ where $z_k = k\,\pi/N$

## Update rule

After observing error $y_t$, the agent updates the value of **all** arc widths:

$$
Q_{t+1}(k) = Q_t(k) + \eta \Big[ \underbrace{(\pi - z_k) \cdot \mathbb{I}(|y_t| \le z_k)}_{\text{counterfactual reward}} - Q_t(k) \Big]
$$

- If $|y_t|$ was small (precise memory), small arcs $z_k \ge |y_t|$ get a large positive signal.
- If $|y_t|$ was large (poor memory), small arcs $z_k < |y_t|$ get a signal of 0 (missed target), driving their value down.

## Action selection (softmax)

The agent selects the next arc width from a softmax policy over Q-values:

$$
P(\alpha_t = z_k \mid Q_t) = \frac{\exp(Q_t(k) / \tau)}{\sum_{j=1}^N \exp(Q_t(j) / \tau)}
$$

where $\tau$ is the temperature parameter — lower values make the policy more greedy (deterministic), higher values more exploratory (uniform).

## Parameters

| Parameter | Description | Range |
|-----------|-------------|-------|
| $\eta$ | Learning rate | $(0, 1]$ |
| $\tau$ | Temperature (softmax) | $(0, \infty)$ |

# Single Run Demonstration

A single run with moderate learning rate and temperature:

```{r}
#| label: crl-single-run
algo_crl(eta = 0.1, tau = 0.05) |>
  run_arc_algorithm(n_trials = 1000, c = 4, init_arc = pi/8) |>
  plot_algorithm_run()

```


# Effect of Learning Rate ($\eta$)

```{r}
#| label: crl-eta-sweep
#| fig-width: 10
#| fig-height: 8
ribbon_summary_runs(
  alg_list = list(
    `eta=0.01` = algo_crl(eta = 0.01, tau = 0.05),
    `eta=0.05` = algo_crl(eta = 0.05, tau = 0.05),
    `eta=0.1`  = algo_crl(eta = 0.1,  tau = 0.05),
    `eta=0.3`  = algo_crl(eta = 0.3,  tau = 0.05)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + theme(legend.position = "bottom")
```

# Effect of Temperature ($\tau$)

Lower $\tau$ makes the policy more deterministic (greedy), concentrating selections around the highest-valued arc width. Higher $\tau$ spreads selections more uniformly, maintaining exploration.

```{r}
#| label: crl-tau-sweep
#| fig-width: 10
#| fig-height: 8
ribbon_summary_runs(
  alg_list = list(
    `tau=1`    = algo_crl(eta = 0.1, tau = 1),
    `tau=0.33` = algo_crl(eta = 0.1, tau = 0.33),
    `tau=0.2`  = algo_crl(eta = 0.1, tau = 0.2),
    `tau=0.1`  = algo_crl(eta = 0.1, tau = 0.1),
    `tau=0.05` = algo_crl(eta = 0.1, tau = 0.05)
  ),
  cs = c(4, 8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + theme(legend.position = "bottom")
```

# Convergence Across Memory Strengths

```{r}
#| label: crl-spaghetti
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(CRL = algo_crl(eta = 0.1, tau = 0.05)),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

# Learned Q-Value Landscape

One of the key advantages of CRL over heuristic algorithms is that it learns a full value landscape $Q(\alpha)$. We can inspect this landscape after training to see whether it captures the true expected reward curve.

```{r}
#| label: q-value-landscape
#| fig-width: 10
#| fig-height: 8
N <- 180
z <- seq(0, pi, length.out = N)

cs_viz <- c(2, 4, 8, 16)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (c_val in cs_viz) {
  state <- new.env(parent = emptyenv())
  state$Q <- rep(0, N)

  y_seq <- rsdm(5000, 0, c = c_val, kappa = 3)

  eta <- 0.05
  for (i in seq_along(y_seq)) {
    cf_reward <- (pi - z) * (abs(y_seq[i]) <= z)
    state$Q <- state$Q + eta * (cf_reward - state$Q)
  }

  true_expected_reward <- sapply(z, \(zk) {
    p_hit <- 2 * psdm(zk, c = c_val, kappa = 3) - 1
    (pi - zk) * p_hit
  })

  q_scaled <- state$Q / max(state$Q) * max(true_expected_reward)

  opt <- find_optimum_sdm(c_val, kappa = 3)

  plot(z, true_expected_reward, type = "l", lwd = 2,
       xlab = expression(alpha), ylab = "Expected reward",
       main = paste0("c = ", c_val))
  lines(z, q_scaled, col = "steelblue", lwd = 2, lty = 2)
  abline(v = opt$alpha, lty = 3, col = "red")
  abline(v = z[which.max(state$Q)], lty = 3, col = "steelblue")
  legend("topright",
         legend = c("True E[reward]", "Learned Q (scaled)", "Optimal", "Q argmax"),
         col = c("black", "steelblue", "red", "steelblue"),
         lty = c(1, 2, 3, 3), lwd = c(2, 2, 1, 1),
         cex = 0.8, bty = "n")
}
```

# Comparison with Algorithm 4 and Algorithm 5

How does CRL compare to the existing heuristic algorithms?

```{r}
#| label: crl-vs-algo4-algo5
#| fig-width: 10
#| fig-height: 8
ribbon_summary_runs(
  alg_list = list(
    algo4      = algo4(delta = 0.1),
    algo5      = algo5(tau = 0.025, delta0 = 0.1),
    CRL        = algo_crl(eta = 0.1, tau = 0.05)
  ),
  cs = c(2, 4, 8),
  n_trials = 200,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + theme(legend.position = "bottom")
```

# Arc Width Distribution

A distinctive property of CRL with softmax action selection is that the stationary arc-width distribution emerges from the softmax over converged Q-values. Let's examine the distribution of chosen arcs after convergence for different parameter settings.

```{r}
#| label: crl-arc-distribution
#| fig-width: 10
#| fig-height: 6
cs_dist <- c(2, 4, 8)
taus_dist <- c(0.33, 0.2, 0.1, 0.05)
pars_dist <- expand.grid(c = cs_dist, tau = taus_dist)

results <- purrr::pmap_dfr(pars_dist, \(c, tau) {
  out <- run_arc_algorithm(algo_crl(eta = 0.1, tau = tau, K = 180),
                           c = c, n_trials = 10000)
  data.frame(c = c, tau = tau, arc = out$a[1001:10000])
})

results |>
  mutate(c = factor(c, labels = paste0("c = ", cs_dist)),
         tau = factor(tau, labels = paste0("tau = ", taus_dist))) |>
  ggplot(aes(x = arc)) +
  geom_histogram(fill = "steelblue", alpha = 0.7) +
  facet_grid(tau ~ c, scales = "free_y") +
  labs(x = "Arc half-width (rad)", y = "Count",
       title = "Stationary arc distribution after burn-in (trials 1001–10000)") +
  theme_minimal()
```


```{r}
#| label: crl-arc-distribution2
#| fig-width: 10
#| fig-height: 6
cs_dist <- c(2, 4, 8)
eta_dist <- c(0.01, 0.025, 0.05, 0.1, 0.2)
pars_dist <- expand.grid(c = cs_dist, eta = eta_dist)

results <- purrr::pmap_dfr(pars_dist, \(c, eta) {
  out <- run_arc_algorithm(algo_crl(eta = eta, tau = 0.05, K = 180),
                           c = c, n_trials = 10000)
  data.frame(c = c, eta = eta, arc = out$a[1001:10000])
})

results |>
  mutate(c = factor(c, labels = paste0("c = ", cs_dist)),
         eta = factor(eta, labels = paste0("eta = ", eta_dist))) |>
  ggplot(aes(x = arc)) +
  geom_histogram(fill = "steelblue", alpha = 0.7) +
  facet_grid(eta ~ c, scales = "free_y") +
  labs(x = "Arc half-width (rad)", y = "Count",
       title = "Stationary arc distribution after burn-in (trials 1001–10000)") +
  theme_minimal()
```

# Summary

**Key properties of Counterfactual Q-Learning:**

1. **Global value map:** Unlike algo4/algo5, CRL learns the full shape of the reward curve $E[r|\alpha]$, giving it a global view of which arc widths are good vs. bad.

2. **Stationary convergence:** Q-values converge to the true expected reward, and the softmax policy settles on a stable distribution centered near the optimum. No oscillation as in algo4 with fixed $\delta$.

3. **Natural stochasticity:** The softmax policy provides built-in stochasticity controlled by $\tau$, without needing separate noise injection. This gives a direct likelihood for fitting to data.

4. **Counterfactual advantage:** Because the agent observes $y_t$ post-hoc, it updates all actions simultaneously. This is far more sample-efficient than updating only the chosen action.

**Considerations for model fitting:**

- The likelihood decomposes naturally: $\mathcal{L} = \sum_t \ln f_{\text{SDM}}(y_t) + \sum_t \ln P(\alpha_t \mid Q_t, \tau)$
- Q-values evolve deterministically from the sequence of observed errors, so no latent state integration is needed
- Free parameters for the arc component: $\eta$ (learning rate) and $\tau$ (temperature)
- Joint with memory: $c$, $\kappa$, $\eta$, $\tau$

## Postscript: how the Q profile changes over trials

```{r}
crl <- algo_crl(eta = 0.1, tau = 0.05)

ys <- c(pi/2, 2, 0.5, 0.1, 3, 1, 0.2)

par(mfrow = c(2,4))
Q <- environment(crl)$state$Q
plot(seq(0, pi, length.out = length(Q)), Q, type = "l", xlab = "Arc width", ylim = c(0,1))

for (i in seq_along(ys)) {
  choice <- crl(a = 2, y = ys[i])
  Q <- environment(crl)$state$Q
  plot(seq(0, pi, length.out = length(Q)), Q, 
  type = "l", xlab = "Arc width", ylim = c(0,1), 
  main = glue::glue("Step {i}, y = {round(ys[i],2)}"))
}



```
