---
title: "Model diagnostics: noisy-gradient (algo5), Bayesian"
format: html
---

```{r cache = FALSE}
#| label: init
#| message: false
library(tidyverse)
library(brms)
library(bmm)
source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: callout-note
## Prerequisites

This notebook fits the noisy-gradient model (algo5) via full Bayesian inference using `brms` and custom Stan code. The same model is fitted by MLE in [algo5 MLE diagnostics](diagnostics-algo5-mle.qmd). The SDM model and optimality derivation are described in [Optimal Arc Width (SDM)](optimal-arc-response.qmd).
:::

## Task and model summary

On each trial a participant reports a remembered color (continuous response) and sets a symmetric **arc** of half-width $a_t \in (0,\pi)$ around it. Points are awarded as 

$$
p_t = (\pi - a_t)\,\mathbf{1}\!\bigl[|y_t| \le a_t\bigr],
$$

where $y_t$ is the signed response error. Wider arcs are safer but earn fewer points, creating an accuracy–confidence tradeoff. [Empirically, participants' arc widths are near-optimal](optimal-arc-response.html) given their error distributions.

**Response errors** are modelled by the [Signal Discrimination Model](https://venpopov.github.io/bmm/articles/bmm_sdm_simple.html) (SDM): 

$$
y_t \sim \text{SDM}(c,\kappa).
$$

**Arc-width updates** follow a noisy stochastic-gradient rule on the logit-reparameterized arc $x_t = \operatorname{logit}(a_t/\pi)$. Define a soft hit indicator and gradient: 

$$
s_t = \sigma\!\left(\frac{a_t - |y_t|}{\tau}\right), \qquad
g_t = -s_t + (\pi - a_t)\,\frac{s_t(1-s_t)}{\tau},
$$ 

with a decaying learning rate $\delta_t = \delta_0/(t+1)^{0.5}$. The state transition is 
$$
x_{t+1} = x_t + \delta_t\bigl(g_t + \varepsilon_t\bigr), \qquad \varepsilon_t \sim \mathcal{N}(0,\sigma_g^2).
$$

**Joint log-likelihood.** The full log-likelihood decomposes as 

$$
\ell(\theta) =
\underbrace{\sum_{t=1}^{T}\log f_{\text{SDM}}(y_t \mid c,\kappa)}_{\text{response errors}}
\;+\;
\underbrace{\sum_{t=1}^{T-1}\log\phi\!\left(x_{t+1};\; x_t + \delta_t\, g_t,\; \delta_t\,\sigma_g\right)}_{\text{arc transitions}},
$$ 

with parameter vector $\theta=(c,\kappa,\tau,\delta_0,\sigma_g)$. To ensure positivity, all parameters are estimated on the log scale.

## Custom brms family and Stan code

We define a `brms` custom family whose log-likelihood is implemented by the `sdm_algo5_lpdf` Stan function in `stan/sdm_algo5_likelihood.stan`. Use log links for all parameters except the fixed intercept `mu` (tan_half). *Important:* **c** is also on

```{r}
#| label: define-family-stanvars
sdm_algo5 <- custom_family(
  "sdm_algo5",
  dpars = c("mu", "c", "kappa", "tau", "delta", "sigmag"),
  links = c("tan_half", "identity", "log", "log", "log", "log"),
  type = "real",
  loop = FALSE,
  vars = c("vreal1", "vreal2", "vreal3")
)

stanvars <- sdm_base_stanvars() +
  stanvar(
    scode = read_file("stan/sdm_algo5_likelihood.stan"),
    block = "functions"
  )
```

## Priors

We use weakly informative priors on the log scale for the algorithm parameters, centered roughly on plausible values ($\tau \approx 0.025 \implies \log \approx -3.7$; $\delta_0 \approx 0.1 \implies \log \approx -2.3$; $\sigma_g \approx 2 \implies \log \approx 0.7$). Also we must ensure initial values for tau and delta are small, otherwise the sampler gets confused.

```{r}
#| label: define-priors
prior <- set_prior("constant(0)", dpar = "mu", class = "Intercept") +
  set_prior("student_t(5, 2, 0.75)", dpar = "c", class = "Intercept") +
  set_prior("student_t(5, 1.75, 0.75)", dpar = "kappa", class = "Intercept") +
  set_prior("normal(-3.5, 2)", dpar = "tau", class = "Intercept") +
  set_prior("normal(-2.5, 2)", dpar = "delta", class = "Intercept") +
  set_prior("normal(0, 2)", dpar = "sigmag", class = "Intercept")

init_algo5_brm <- function() {
  tau_init <- log(runif(1, 1e-6, 0.05))
  delta_init <- log(runif(1, 1e-6, 0.5))
  list(
    Intercept_mu = 0,
    Intercept_c = runif(1, -2, 2),
    Intercept_kappa = runif(1, -2, 2),
    Intercept_tau = tau_init,
    Intercept_delta = delta_init,
    Intercept_sigmag = runif(1, -2, 2)
  )
}
```

## Simulate data

Generate a single subject's worth of data from the algo5 model with known parameters.

```{r}
#| label: simulate-data
true_pars <- list(c = 6, kappa = 3, tau = 0.025, delta = 0.1, sigma_g = 2)

set.seed(4233)
sim <- run_arc_algorithm(
  arc_update_fun = algo5_noisy_grad(
    tau = true_pars$tau,
    delta0 = true_pars$delta,
    delta_pow = 0.5,
    sigma_g = true_pars$sigma_g
  ),
  c = true_pars$c,
  kappa = true_pars$kappa,
  init_arc = pi / 8,
  n_trials = 1000
)

sim_data <- tibble(y = sim$y, a = sim$a) |>
  mutate(a_next = lead(a, default = -1), trial = row_number())

par(mfrow = c(1, 3))
plot(sim$a, type = "l", xlab = "Trial", ylab = "Arc width")
hist(sim$a, xlab = "Arc width")
hist(sim$y, xlab = "Response error")
par(mfrow = c(1, 1))
```

## Fit simulated data with brms

We fit the model using `brms` with the custom family.

```{r}
#| label: fit-sim
fit_sim <- brm(
  bf(y | vreal(a, a_next, trial) ~ 1, c ~ 1, kappa ~ 1, tau ~ 1, delta ~ 1, sigmag ~ 1),
  prior = prior,
  data = sim_data, family = sdm_algo5, stanvars = stanvars,
  backend = "cmdstanr", cores = 4, init = init_algo5_brm,
  file = "output/algo5_bayes_sim_fit.rds", file_refit = "on_change"
)
```

## Diagnostics

```{r}
#| label: fit-summary
summary(fit_sim)
```

```{r}
#| label: fit-plot
#| fig-width: 10
#| fig-height: 6
plot(fit_sim)
```

## Compare estimated vs true parameters

Extract posterior means and compare to generating values. Note all parameters use log link.

```{r}
#| label: compare-params
sdm5_pars_from_fit <- function(fit) {
  est <- fixef(fit)[-1, 1]
  list(
    c = exp(est["c_Intercept"]),
    kappa = exp(est["kappa_Intercept"]),
    tau = exp(est["tau_Intercept"]),
    delta = exp(est["delta_Intercept"]),
    sigma_g = exp(est["sigmag_Intercept"])
  )
}

estimated <- sdm5_pars_from_fit(fit_sim)

comparison <- tibble(
  parameter = names(true_pars),
  true_value = unlist(true_pars),
  estimated = unlist(estimated)
)
comparison
```

## Posterior predictive check: simulated arc trajectories

Simulate arc trajectories from the posterior to check whether the fitted model reproduces the observed arc dynamics.

```{r}
#| label: pp-arc-trajectories
#| fig-width: 10
#| fig-height: 5
est <- estimated
sim_pp <- run_arc_algorithm(
  arc_update_fun = algo5_noisy_grad(
    tau = est$tau,
    delta0 = est$delta,
    delta_pow = 0.5,
    sigma_g = est$sigma_g
  ),
  c = est$c,
  kappa = est$kappa,
  init_arc = sim$a[1],
  n_trials = nrow(sim_data)
)

par(mfrow = c(1, 2))
plot(sim$a, type = "l", xlab = "Trial", ylab = "Arc width", main = "Arc trajectory")
lines(sim_pp$a, col = "red")
legend("topright", c("Data", "Posterior"), col = c("black", "red"), lty = 1)

plot(density(sim$a), main = "Arc width distribution")
lines(density(sim_pp$a), col = "red")
legend("topright", c("Data", "Posterior"), col = c("black", "red"), lty = 1)
par(mfrow = c(1, 1))
```

## Parameter recovery

Assess parameter recovery for the Bayesian version by simulating many datasets, fitting each with `brms`. Spot check with just a 100 datasets (~ 17 minutes) - the [MLE recovery diagnostics](diagnostics-algo5-mle.qmd) tested it for 10000 and showed great recovery. The likelihood is the same, so I expect similar performance, but the bayesian fits are slower. So this is a sanity check.

```{r}
#| label: param-recovery-bayes
n_recovery <- 100
n_trials <- 1000
set.seed(2026)

recovery_true <- tibble(
  iter = seq_len(n_recovery),
  c = runif(n_recovery, 1, 10),
  kappa = runif(n_recovery, 1, 10),
  tau = runif(n_recovery, 0.001, 0.05),
  delta0 = runif(n_recovery, 0.01, 0.2),
  sigma_g = runif(n_recovery, 0, 10)
)

fit_one_bayes <- function(iter, c, kappa, tau, delta0, sigma_g) {
  sim <- run_arc_algorithm(
    arc_update_fun = algo5_noisy_grad(
      tau = tau, delta0 = delta0, delta_pow = 0.5, sigma_g = sigma_g
    ),
    c = c, kappa = kappa, init_arc = pi / 2, n_trials = n_trials
  )
  sim_data <- tibble(y = sim$y, a = sim$a) |>
    mutate(a_next = lead(a, default = -1), trial = row_number())
  
  fit <- tryCatch(
    brm(
      bf(y | vreal(a, a_next, trial) ~ 1, c ~ 1, kappa ~ 1, tau ~ 1, delta ~ 1, sigmag ~ 1),
      prior = prior,
      data = sim_data, family = sdm_algo5, stanvars = stanvars,
      backend = "cmdstanr", cores = 4, init = init_algo5_brm,
      refresh = 0, silent = 2
    ),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    return(tibble(
      iter = iter, c_est = NA, kappa_est = NA, tau_est = NA,
      delta_est = NA, sigma_g_est = NA,
      rhat_max = NA, converged = FALSE
    ))
  }
  est <- tryCatch({
    s <- fixef(fit)[-1, 1]
    list(
      c = exp(s["c_Intercept"]),
      kappa = exp(s["kappa_Intercept"]),
      tau = exp(s["tau_Intercept"]),
      delta = exp(s["delta_Intercept"]),
      sigma_g = exp(s["sigmag_Intercept"])
    )
  }, error = function(e) list(c = NA, kappa = NA, tau = NA, delta = NA, sigma_g = NA))
  
  rhat_max <- max(posterior::rhat(fit), na.rm = TRUE)
  tibble(
    iter = iter,
    c_est = est$c, kappa_est = est$kappa,
    tau_est = est$tau, delta_est = est$delta, sigma_g_est = est$sigma_g,
    rhat_max = rhat_max, converged = (rhat_max < 1.1)
  )
}

recovery_est_bayes <- with_cache(
  "output/algo5_recovery_bayes.rds",
  purrr::pmap_dfr(recovery_true, fit_one_bayes)
)
```

Tag iterations with poor convergence or extreme estimates:

```{r}
#| label: tag-recovery-outliers
est_cols <- c("c_est", "kappa_est", "tau_est", "delta_est", "sigma_g_est")
true_cols <- c("c", "kappa", "tau", "delta0", "sigma_g")
thresholds <- 4 * sapply(true_cols, function(col) max(recovery_true[[col]], na.rm = TRUE))
outlier_mat <- sapply(seq_along(est_cols), function(j) {
  recovery_est_bayes[[est_cols[j]]] > thresholds[j]
})
recovery_est_bayes$bad <- rowSums(outlier_mat, na.rm = TRUE) > 0 | !recovery_est_bayes$converged
```

Join true and estimated values and reshape for plotting:

```{r}
#| label: build-recovery-df
recovery_df_bayes <- recovery_true |>
  left_join(recovery_est_bayes, by = "iter") |>
  pivot_longer(
    cols = c(c, kappa, tau, delta0, sigma_g),
    names_to = "parameter",
    values_to = "true_value"
  ) |>
  mutate(
    estimated = case_when(
      parameter == "c" ~ c_est,
      parameter == "kappa" ~ kappa_est,
      parameter == "tau" ~ tau_est,
      parameter == "delta0" ~ delta_est,
      parameter == "sigma_g" ~ sigma_g_est
    )
  ) |>
  select(iter, parameter, true_value, estimated, converged, bad)

cat("Converged:", sum(recovery_est_bayes$converged, na.rm = TRUE), "/", n_recovery,
    sprintf("(%.1f%%)", 100 * mean(recovery_est_bayes$converged, na.rm = TRUE)), "\n")
```

Correlations between true and estimated values (converged runs only):

```{r}
#| label: recovery-correlations
recovery_df_bayes |>
  filter(converged, !bad) |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated, use = "complete.obs"),
    rmse = sqrt(mean((true_value - estimated)^2, na.rm = TRUE)),
    bias = mean(estimated - true_value, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery-bayes
#| fig-cap: "Parameter recovery for the noisy-gradient model (Bayesian)."

plot_df <- recovery_df_bayes |> filter(converged, !bad)

cor_labels <- plot_df |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated, use = "complete.obs"),
    x = -Inf, y = Inf,
    .groups = "drop"
  ) |>
  mutate(label = sprintf("r = %.2f", r))

plot_df |>
  ggplot(aes(x = true_value, y = estimated)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_point(aes(color = converged), alpha = 0.5, size = 1.5) +
  geom_text(data = cor_labels, aes(x = x, y = y, label = label),
            hjust = -0.1, vjust = 1.3, inherit.aes = FALSE, size = 4) +
  facet_wrap(~parameter, scales = "free") +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "red")) +
  labs(x = "True value", y = "Estimated value", color = "Converged")
```

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery-binned-bayes
#| fig-cap: "Binned parameter recovery (Bayesian)."

binned_df <- plot_df |>
  group_by(parameter) |>
  mutate(bin = cut(true_value, breaks = 20, labels = FALSE)) |>
  group_by(parameter, bin) |>
  summarise(
    true_mid = mean(true_value, na.rm = TRUE),
    median_est = median(estimated, na.rm = TRUE),
    q025 = quantile(estimated, 0.025, na.rm = TRUE),
    q10 = quantile(estimated, 0.10, na.rm = TRUE),
    q90 = quantile(estimated, 0.90, na.rm = TRUE),
    q975 = quantile(estimated, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(binned_df, aes(x = true_mid)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = q10, ymax = q90), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = median_est), color = "steelblue", linewidth = 0.8) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "True value", y = "Estimated value")
```

## Hierarchical recovery

Simulate 100 subjects whose parameters are drawn from population-level
distributions centered on the true parameters defined earlier. Random effects
are drawn on the linear predictor scale (log for all algo5 parameters except c
which uses identity), then transformed to natural scale for simulation.

```{r}
#| label: hier-recovery-simulate
n_subjects <- 100
n_trials_per_subj <- 1000

pop_intercepts <- list(
  c = log(true_pars$c),
  kappa = log(true_pars$kappa),
  tau = log(true_pars$tau),
  delta = log(true_pars$delta),
  sigmag = log(true_pars$sigma_g)
)

re_sds_true <- list(c = 0.2, kappa = 0.3, tau = 0.3, delta = 0.3, sigmag = 0.3)

set.seed(2026)
true_re <- tibble(
  subject = seq_len(n_subjects),
  re_c = rnorm(n_subjects, 0, re_sds_true$c),
  re_kappa = rnorm(n_subjects, 0, re_sds_true$kappa),
  re_tau = rnorm(n_subjects, 0, re_sds_true$tau),
  re_delta = rnorm(n_subjects, 0, re_sds_true$delta),
  re_sigmag = rnorm(n_subjects, 0, re_sds_true$sigmag)
) |>
  mutate(
    c = exp(pop_intercepts$c + re_c),
    kappa = exp(pop_intercepts$kappa + re_kappa),
    tau = exp(pop_intercepts$tau + re_tau),
    delta = exp(pop_intercepts$delta + re_delta),
    sigma_g = exp(pop_intercepts$sigmag + re_sigmag)
  )

hier_sim_data <- pmap_dfr(
  true_re,
  \(subject, c, kappa, tau, delta, sigma_g, ...) {
    sim <- run_arc_algorithm(
      arc_update_fun = algo5_noisy_grad(
        tau = tau, delta0 = delta, delta_pow = 0.5, sigma_g = sigma_g
      ),
      c = c, kappa = kappa,
      init_arc = pi / 4, n_trials = n_trials_per_subj
    )
    tibble(subject = subject, y = sim$y, a = sim$a)
  }
) |>
  group_by(subject) |>
  mutate(a_next = lead(a, default = -1), trial = row_number()) |>
  ungroup()
```

### Fit the hierarchical model

```{r}
#| label: hier-recovery-fit
hier_fit <- brm(
    bf(
      y | vreal(a, a_next, trial) ~ 1,
      c ~ 1 + (1 | subject),
      kappa ~ 1 + (1 | subject),
      tau ~ 1 + (1 | subject),
      delta ~ 1 + (1 | subject),
      sigmag ~ 1 + (1 | subject)
    ),
    prior = prior,
    data = hier_sim_data, family = sdm_algo5, stanvars = stanvars,
    backend = "cmdstanr", cores = 4, init = init_algo5_brm, 
    file = "output/algo5_bayes_hier_recovery.rds",
    refresh = 0
  )
```

Here's a diagnostic plot:

```{r}
plot(hier_fit, ask = FALSE)
```

Unfortunately this took an extremely long time because one chain got very confused. 

::: callout-caution

All the results below are not to be trusted due to the divergent chain found above

:::

```{r}
#| label: hier-recovery-summary
summary(hier_fit)
```

### Compare population-level parameters

```{r}
#| label: hier-recovery-fixef
fe <- fixef(hier_fit)
pop_comparison <- tibble(
  parameter = c("c", "kappa", "tau", "delta", "sigmag"),
  true_intercept = unlist(pop_intercepts),
  est_intercept = fe[c("c_Intercept", "kappa_Intercept", "tau_Intercept",
                       "delta_Intercept", "sigmag_Intercept"), "Estimate"]
)
pop_comparison
```

Compare estimated group-level SDs to the true generating SDs:

```{r}
#| label: hier-recovery-sd
sd_est <- VarCorr(hier_fit)$subject$sd
sd_comparison <- tibble(
  parameter = c("c", "kappa", "tau", "delta", "sigmag"),
  true_sd = unlist(re_sds_true),
  est_sd = sd_est[, "Estimate"]
)
sd_comparison
```

### Compare subject-level random effects

```{r}
#| label: hier-recovery-ranef
re <- ranef(hier_fit)$subject

est_re <- tibble(
  subject = seq_len(n_subjects),
  re_c_est = re[, "Estimate", "c_Intercept"],
  re_kappa_est = re[, "Estimate", "kappa_Intercept"],
  re_tau_est = re[, "Estimate", "tau_Intercept"],
  re_delta_est = re[, "Estimate", "delta_Intercept"],
  re_sigmag_est = re[, "Estimate", "sigmag_Intercept"]
)

compare_re <- true_re |>
  select(subject, re_c, re_kappa, re_tau, re_delta, re_sigmag) |>
  left_join(est_re, by = "subject")

re_long <- bind_rows(
  tibble(parameter = "c", true = compare_re$re_c, estimated = compare_re$re_c_est),
  tibble(parameter = "kappa", true = compare_re$re_kappa, estimated = compare_re$re_kappa_est),
  tibble(parameter = "tau", true = compare_re$re_tau, estimated = compare_re$re_tau_est),
  tibble(parameter = "delta", true = compare_re$re_delta, estimated = compare_re$re_delta_est),
  tibble(parameter = "sigmag", true = compare_re$re_sigmag, estimated = compare_re$re_sigmag_est)
)
```

```{r}
#| label: hier-recovery-re-cor
re_long |>
  group_by(parameter) |>
  summarise(
    r = cor(true, estimated, use = "complete.obs"),
    rmse = sqrt(mean((true - estimated)^2, na.rm = TRUE)),
    bias = mean(estimated - true, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| fig-width: 10
#| fig-height: 10
#| label: fig-hier-recovery-re
#| fig-cap: "Hierarchical recovery of subject-level random effects (linear predictor scale). Each point is one subject. The dashed line is the identity."

cor_labels <- re_long |>
  group_by(parameter) |>
  summarise(
    r = cor(true, estimated, use = "complete.obs"),
    x = -Inf, y = Inf,
    .groups = "drop"
  ) |>
  mutate(label = sprintf("r = %.2f", r))

ggplot(re_long, aes(x = true, y = estimated)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_text(data = cor_labels, aes(x = x, y = y, label = label),
            hjust = -0.1, vjust = 1.3, inherit.aes = FALSE, size = 4) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "True random effect", y = "Estimated random effect")
```
