---
title: "Model diagnostics: contraction-expansion (algo4) with logit noise, Bayesian"
format: html
---

```{r cache = FALSE}
#| label: setup
#| message: false
library(tidyverse)
library(brms)
library(bmm)
source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: {.callout-note}
## Prerequisites
This notebook fits the contraction-expansion model (algo4) with logit-scale noise
via full Bayesian inference using `brms` and custom Stan code. The same model is
fitted by MLE in [algo4 MLE diagnostics](diagnostics-algo4-mle.qmd).
The SDM model and optimality derivation are described in
[Optimal Arc Width (SDM)](optimal-arc-response.qmd).
:::

## Task and model summary

On each trial a participant reports a remembered color (continuous response) and
sets a symmetric **arc** of half-width $a_t \in (0,\pi)$ around it. Points are
awarded as
$$
p_t = (\pi - a_t)\,\mathbf{1}\!\bigl[|y_t| \le a_t\bigr],
$$
where $y_t$ is the signed response error. Wider arcs are safer but earn fewer
points, creating an accuracy–confidence tradeoff.
[Empirically, participants' arc widths are near-optimal](optimal-arc-response.html)
given their error distributions.

**Response errors** are modelled by the
[Signal Discrimination Model](https://venpopov.github.io/bmm/articles/bmm_sdm_simple.html)
(SDM):
$$
y_t \sim \text{SDM}(c,\kappa).
$$

**Deterministic arc-width update (algo4).** On each trial the arc contracts
toward 0 on a hit and expands toward $\pi$ on a miss, with step-size
$\delta \in (0,1)$:
$$
\tilde{a}_{t+1} = a_t\,(1-\delta) + \mathbf{1}\!\bigl[|y_t|>a_t\bigr]\,\delta\,\pi.
$$

**Logit-noise stochastic extension.** To make the model estimable, we add
Gaussian noise on the unconstrained logit scale. Define
$$
\tilde{x}_{t+1} = \operatorname{logit}\!\left(\tilde{a}_{t+1}/\pi\right),
$$
then the stochastic transition is
$$
x_{t+1} = \tilde{x}_{t+1} + \varepsilon_t, \qquad
\varepsilon_t \sim \mathcal{N}(0,\sigma_\varepsilon^2),
$$
so that $a_{t+1} = \pi\,\sigma(x_{t+1})$ is guaranteed to lie in $(0,\pi)$
without clipping.

**Joint log-likelihood.** Because $x_{t+1} \mid a_t, y_t$ is normal, the full
log-likelihood decomposes as
$$
\ell(\theta) =
\underbrace{\sum_{t=1}^{T}\log f_{\text{SDM}}(y_t \mid c,\kappa)}_{\text{response errors}}
\;+\;
\underbrace{\sum_{t=1}^{T-1}\log\phi\!\left(x_{t+1};\; \tilde{x}_{t+1},\; \sigma_\varepsilon\right)}_{\text{arc transitions}},
$$
with parameter vector $\theta=(c,\kappa,\delta,\sigma_\varepsilon)$.

## Custom brms family and Stan code

We define a `brms` custom family whose log-likelihood is implemented by the
`sdm_algo4_lpdf` Stan function in `stan/sdm_algo4_likelihood.stan`. The SDM
numerator and normalization are provided by `bmm`'s Stan chunks.

```{r}
#| label: define-family-stanvars
sdm_algo4 <- custom_family(
  "sdm_algo4",
  dpars = c("mu", "c", "kappa", "delta", "asigma"),
  links = c("tan_half", "identity", "log", "logit", "log"),
  type = "real",
  loop = FALSE,
  vars = "vreal1"
)

stanvars <- sdm_base_stanvars() +
  stanvar(
    scode = read_file("stan/sdm_algo4_likelihood.stan"),
    block = "functions"
  )
```

## Priors

For c and kappa we use the default bmm priors. For delta and asigma we use a weakly informative prior that has more mass close to 0 but allows potentially larger values for normalization.

```{r}
#| label: define-priors
prior <- set_prior("constant(0)", dpar = "mu", class = "Intercept") +
  set_prior("student_t(5, 2, 0.75)", dpar = "c", class = "Intercept") +
  set_prior("student_t(5, 1.75, 0.75)", dpar = "kappa", class = "Intercept") +
  set_prior("normal(-3.5, 2)", dpar = "delta", class = "Intercept") +
  set_prior("normal(-3.5, 2)", dpar = "asigma", class = "Intercept")
```

## Simulate data

Generate a single subject's worth of data from the algo4 logit-noise model with
known parameters.

```{r}
#| label: simulate-data
true_pars <- list(c = 6, kappa = 3, delta = 0.1, sigma_e = 0.2)

set.seed(4233)
sim <- run_arc_algorithm(
  arc_update_fun = algo4_logit_noise(
    delta = true_pars$delta,
    sigma_e = true_pars$sigma_e
  ),
  c = true_pars$c,
  kappa = true_pars$kappa,
  init_arc = pi / 8,
  n_trials = 1000
)

sim_data <- tibble(y = sim$y, a = sim$a)

par(mfrow = c(1, 3))
plot(sim$a, type = "l", xlab = "Trial", ylab = "Arc width")
hist(sim$a, xlab = "Arc width")
hist(sim$y, xlab = "Response error")
par(mfrow = c(1, 1))
```

## Fit simulated data with brms

We fit the model using `brms` with the custom family and Stan code defined
above. The `vreal(a)` syntax passes the arc-width vector as additional data to
the Stan likelihood.

```{r}
#| label: fit-sim
fit_sim <- brm(
  bf(y | vreal(a) ~ 1, c ~ 1, kappa ~ 1, delta ~ 1, asigma ~ 1),
  prior = prior,
  data = sim_data, family = sdm_algo4, stanvars = stanvars,
  backend = "cmdstanr", cores = 4, init = 0,
  file = "output/algo4_bayes_sim_fit.rds", file_refit = "on_change"
)
```

## Diagnostics

```{r}
#| label: fit-summary
summary(fit_sim)
```

```{r}
#| label: fit-plot
#| fig-width: 10
#| fig-height: 6
plot(fit_sim)
```

## Compare estimated vs true parameters

Extract posterior means and compare to generating values.

```{r}
#| label: compare-params
sdm4_pars_from_fit <- function(fit) {
  est <- fixef(fit)[-1, 1]
  list(
    c = exp(est["c_Intercept"]),
    kappa = exp(est["kappa_Intercept"]),
    delta = inv_logit(est["delta_Intercept"]),
    sigma_e = exp(est["asigma_Intercept"])
  )
}

estimated <- sdm4_pars_from_fit(fit_sim)

comparison <- tibble(
  parameter = names(true_pars),
  true_value = unlist(true_pars),
  estimated = unlist(estimated)
)
comparison
```

## Posterior predictive check: simulated arc trajectories

Simulate arc trajectories from the posterior to check whether the fitted model
reproduces the observed arc dynamics.

```{r}
#| label: pp-arc-trajectories
#| fig-width: 10
#| fig-height: 5
est <- estimated
sim_pp <- run_arc_algorithm(
  arc_update_fun = algo4_logit_noise(
    delta = est$delta,
    sigma_e = est$sigma_e
  ),
  c = est$c,
  kappa = est$kappa,
  init_arc = sim$a[1],
  n_trials = nrow(sim_data)
)

par(mfrow = c(1, 2))
plot(sim$a, type = "l", xlab = "Trial", ylab = "Arc width", main = "Arc trajectory")
lines(sim_pp$a, col = "red")
legend("topright", c("Data", "Posterior"), col = c("black", "red"), lty = 1)

plot(density(sim$a), main = "Arc width distribution")
lines(density(sim_pp$a), col = "red")
legend("topright", c("Data", "Posterior"), col = c("black", "red"), lty = 1)
par(mfrow = c(1, 1))
```

## Parameter recovery

Assess parameter recovery for the Bayesian version by simulating many datasets, fitting each with `brms`, and extracting only summary statistics and convergence diagnostics (not saving full fits).

```{r}
#| label: param-recovery-bayes
n_recovery <- 1000  
n_trials <- 1000
set.seed(2026)

recovery_true <- tibble(
  iter = seq_len(n_recovery),
  c = runif(n_recovery, 1, 10),
  kappa = runif(n_recovery, 1, 10),
  delta = runif(n_recovery, 0.01, 0.5),
  sigma_e = runif(n_recovery, 0.01, 1)
)

fit_one_bayes <- function(iter, c, kappa, delta, sigma_e) {
  sim <- run_arc_algorithm(
    arc_update_fun = algo4_logit_noise(delta = delta, sigma_e = sigma_e),
    c = c, kappa = kappa, init_arc = pi / 2, n_trials = n_trials
  )
  sim_data <- tibble(y = sim$y, a = sim$a)
  
  fit <- tryCatch(
    brm(
      bf(y | vreal(a) ~ 1, c ~ 1, kappa ~ 1, delta ~ 1, asigma ~ 1),
      prior = prior,
      data = sim_data, family = sdm_algo4, stanvars = stanvars,
      backend = "cmdstanr", cores = 4,
      refresh = 0, silent = 2
    ),
    error = function(e) NULL
  )
  if (is.null(fit)) {
    return(tibble(
      iter = iter, c_est = NA, kappa_est = NA, delta_est = NA, sigma_e_est = NA,
      rhat_max = NA, neff_min = NA, converged = FALSE
    ))
  }
  est <- tryCatch({
    s <- fixef(fit)[-1, 1]
    list(
      c = exp(s["c_Intercept"]),
      kappa = exp(s["kappa_Intercept"]),
      delta = inv_logit(s["delta_Intercept"]),
      sigma_e = exp(s["asigma_Intercept"])
    )
  }, error = function(e) list(c = NA, kappa = NA, delta = NA, sigma_e = NA))
  rhat_max <- max(posterior::rhat(fit), na.rm = TRUE)
  tibble(
    iter = iter,
    c_est = est$c, kappa_est = est$kappa,
    delta_est = est$delta, sigma_e_est = est$sigma_e,
    rhat_max = rhat_max, converged = (rhat_max < 1.1)
  )
}

recovery_est_bayes <- with_cache(
  "output/algo4_recovery_bayes.rds",
  purrr::pmap_dfr(recovery_true, fit_one_bayes, .progress = TRUE)
)
```

Tag iterations with poor convergence or extreme estimates:

```{r}
#| label: tag-recovery-outliers
est_cols <- c("c_est", "kappa_est", "delta_est", "sigma_e_est")
true_cols <- c("c", "kappa", "delta", "sigma_e")
thresholds <- 4 * sapply(true_cols, function(col) max(recovery_true[[col]], na.rm = TRUE))
outlier_mat <- sapply(seq_along(est_cols), function(j) {
  recovery_est_bayes[[est_cols[j]]] > thresholds[j]
})
recovery_est_bayes$bad <- rowSums(outlier_mat, na.rm = TRUE) > 0 | !recovery_est_bayes$converged
```

Join true and estimated values and reshape for plotting:

```{r}
#| label: build-recovery-df
recovery_df_bayes <- recovery_true |>
  left_join(recovery_est_bayes, by = "iter") |>
  pivot_longer(
    cols = c(c, kappa, delta, sigma_e),
    names_to = "parameter",
    values_to = "true_value"
  ) |>
  mutate(
    estimated = case_when(
      parameter == "c" ~ c_est,
      parameter == "kappa" ~ kappa_est,
      parameter == "delta" ~ delta_est,
      parameter == "sigma_e" ~ sigma_e_est
    )
  ) |>
  select(iter, parameter, true_value, estimated, converged, bad)

cat("Converged:", sum(recovery_est_bayes$converged, na.rm = TRUE), "/", n_recovery,
    sprintf("(%.1f%%)", 100 * mean(recovery_est_bayes$converged, na.rm = TRUE)), "\n")
cat("Bad parameter estimates:", sum(recovery_est_bayes$bad, na.rm = TRUE), "/", n_recovery,
    sprintf("(%.1f%%)", 100 * mean(recovery_est_bayes$bad, na.rm = TRUE)), "\n")
```

Correlations between true and estimated values (converged runs only):

```{r}
#| label: recovery-correlations
recovery_df_bayes |>
  filter(converged, !bad) |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated, use = "complete.obs"),
    rmse = sqrt(mean((true_value - estimated)^2, na.rm = TRUE)),
    bias = mean(estimated - true_value, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery-bayes
#| fig-cap: "Parameter recovery for the contraction-expansion model with logit noise (Bayesian). Each point is one simulated dataset. The dashed line is the identity."

plot_df <- recovery_df_bayes |> filter(converged, !bad)

cor_labels <- plot_df |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated, use = "complete.obs"),
    x = -Inf, y = Inf,
    .groups = "drop"
  ) |>
  mutate(label = sprintf("r = %.2f", r))

plot_df |>
  ggplot(aes(x = true_value, y = estimated)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_point(aes(color = converged), alpha = 0.5, size = 1.5) +
  geom_text(data = cor_labels, aes(x = x, y = y, label = label),
            hjust = -0.1, vjust = 1.3, inherit.aes = FALSE, size = 4) +
  facet_wrap(~parameter, scales = "free") +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "red")) +
  labs(x = "True value", y = "Estimated value", color = "Converged")
```

Binned summary: for each parameter, true values are split into 20 equal-width bins. Within each bin, we show the median estimate (line) and 80%/95% intervals (ribbons).

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery-binned-bayes
#| fig-cap: "Binned parameter recovery (Bayesian). Line = median estimate per bin; dark ribbon = 80% interval; light ribbon = 95% interval."

binned_df <- plot_df |>
  group_by(parameter) |>
  mutate(bin = cut(true_value, breaks = 20, labels = FALSE)) |>
  group_by(parameter, bin) |>
  summarise(
    true_mid = mean(true_value, na.rm = TRUE),
    median_est = median(estimated, na.rm = TRUE),
    q025 = quantile(estimated, 0.025, na.rm = TRUE),
    q10 = quantile(estimated, 0.10, na.rm = TRUE),
    q90 = quantile(estimated, 0.90, na.rm = TRUE),
    q975 = quantile(estimated, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(binned_df, aes(x = true_mid)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = q10, ymax = q90), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = median_est), color = "steelblue", linewidth = 0.8) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "True value", y = "Estimated value")
```

