---
title: "Model diagnostics: Counterfactual Q-Learning (CRL), MLE"
format: html
---

```{r cache = FALSE}
#| label: init
#| message: false
library(tidyverse)
library(bmm)
source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: {.callout-note}
## Prerequisites
This notebook fits the Counterfactual Q-Learning model (CRL), developed in [Counterfactual RL](counterfactual-rl.qmd). The SDM model and optimality derivation are described in [Optimal Arc Width (SDM)](optimal-arc-response.qmd).
:::

## Task and model summary

On each trial a participant reports a remembered color and sets a symmetric **arc** of half-width $a_t \in (0,\pi)$. Points are awarded as $p_t = (\pi - a_t)\,\mathbf{1}[|y_t| \le a_t]$. **Response errors** follow the SDM: $y_t \sim \text{SDM}(c,\kappa)$.

**Counterfactual Q-Learning.** The agent discretizes $[0,\pi]$ into $K$ bins $z_1,\dots,z_K$ and maintains Q-values updated on every trial using the counterfactual reward for *all* actions:
$$
Q_{t+1}(k) = Q_t(k) + \eta\bigl[(\pi - z_k)\,\mathbf{1}[|y_t| \le z_k] - Q_t(k)\bigr].
$$

Action selection uses a softmax policy:
$$
P(\alpha_t = z_k \mid Q_t) = \frac{\exp(Q_t(k)/\tau)}{\sum_j \exp(Q_t(j)/\tau)}.
$$

**Joint log-likelihood.** Because Q-values evolve deterministically from observed errors:
$$
\ell(\theta) = \sum_t \log f_{\text{SDM}}(y_t \mid c,\kappa) + \sum_t \log P(\alpha_t \mid Q_t, \tau).
$$

Parameters: $\theta = (c, \kappa, \eta, \tau)$.

## Simulate data

```{r}
#| label: simulate-data
true_pars <- list(c = 6, kappa = 3, eta = 0.1, tau = 0.05)

set.seed(4233)
sim <- run_arc_algorithm(
  arc_update_fun = algo_crl(eta = true_pars$eta, tau = true_pars$tau),
  c = true_pars$c,
  kappa = true_pars$kappa,
  init_arc = pi / 2,
  n_trials = 1000
)

par(mfrow = c(1, 3))
plot(sim$a, type = "l", xlab = "Trial", ylab = "Arc width")
hist(sim$a, xlab = "Arc width")
hist(sim$y, xlab = "Response error")
par(mfrow = c(1, 1))
```

## Fit with `optim`

```{r}
#| label: fit-optim
init_pars <- list(c = 3, kappa = 5, eta = 0.05, tau = 0.1)
theta0 <- par_to_unconstrained_crl(init_pars)

fit <- optim(
  par = theta0,
  fn = neg_loglik_crl,
  y = sim$y,
  a = sim$a,
  method = "Nelder-Mead",
  control = list(maxit = 10000, reltol = 1e-10)
)

fit$convergence
```

## Compare estimated vs true parameters

```{r}
#| label: compare-params
estimated <- par_from_unconstrained_crl(fit$par)

comparison <- tibble(
  parameter = names(true_pars),
  initial = unlist(init_pars),
  true_value = unlist(true_pars),
  estimated = unlist(estimated)
)
comparison
```

## Likelihood profiles at the MLE

```{r}
#| label: likelihood-profile
#| fig-width: 9
#| fig-height: 6
best_theta <- fit$par
par_names <- c("c", "kappa", "eta", "tau")

profiles <- purrr::map_dfr(seq_along(par_names), \(j) {
  grid <- seq(best_theta[j] - 1.5, best_theta[j] + 1.5, length.out = 60)
  ll_vals <- sapply(grid, \(v) {
    theta_tmp <- best_theta
    theta_tmp[j] <- v
    -neg_loglik_crl(theta_tmp, y = sim$y, a = sim$a)
  })

  if (par_names[j] == "eta") {
    value <- inv_logit(grid)
  } else {
    value <- exp(grid)
  }

  tibble(
    parameter = par_names[j],
    value = value,
    loglik = ll_vals
  )
})

ggplot(profiles, aes(x = value, y = loglik)) +
  geom_line() +
  geom_vline(data = comparison, aes(xintercept = estimated),
             linetype = "dashed", color = "blue") +
  geom_vline(data = comparison, aes(xintercept = true_value),
             linetype = "dotted", color = "red") +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "Parameter value", y = "Log-likelihood",
       caption = "Blue dashed = MLE, Red dotted = true value")
```

## Sanity-check recovery across a few parameter sets

Fit a handful of parameter combinations to verify recovery without running a full grid.

```{r}
#| label: sanity-recovery
recovery_sets <- tibble(
  c     = c(2, 4, 8, 6, 10, 3),
  kappa = c(3, 3, 3, 5, 2,  8),
  eta   = c(0.05, 0.1, 0.2, 0.15, 0.1, 0.05),
  tau   = c(0.1, 0.05, 0.05, 0.1, 0.02, 0.2)
)

set.seed(2026)

fit_one_crl <- function(c, kappa, eta, tau, n_trials = 1000) {
  sim <- run_arc_algorithm(
    arc_update_fun = algo_crl(eta = eta, tau = tau),
    c = c, kappa = kappa, init_arc = pi / 2, n_trials = n_trials
  )

  theta0 <- par_to_unconstrained_crl(list(
    c = 4, kappa = 4, eta = 0.1, tau = 0.1
  ))

  fit <- tryCatch(
    optim(par = theta0, fn = neg_loglik_crl, y = sim$y, a = sim$a,
          method = "Nelder-Mead",
          control = list(maxit = 10000, reltol = 1e-10)),
    error = \(e) list(par = rep(NA_real_, 4), convergence = 99)
  )

  est <- par_from_unconstrained_crl(fit$par)
  tibble(
    c_true = c, kappa_true = kappa, eta_true = eta, tau_true = tau,
    c_est = est$c, kappa_est = est$kappa, eta_est = est$eta, tau_est = est$tau,
    converged = fit$convergence == 0
  )
}

recovery_results <- purrr::pmap_dfr(recovery_sets, fit_one_crl)
```

```{r}
#| label: recovery-table
recovery_long <- recovery_results |>
  mutate(set = row_number()) |>
  pivot_longer(
    cols = -c(set, converged),
    names_to = c("parameter", ".value"),
    names_pattern = "(.+)_(true|est)"
  )

recovery_long |>
  mutate(error = est - true, rel_error = error / true) |>
  select(set, parameter, true, est, error, rel_error, converged)
```

```{r}
#| label: fig-sanity-recovery
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Sanity-check parameter recovery for CRL. Each point is one parameter set."
recovery_long |>
  ggplot(aes(x = true, y = est)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_point(size = 3, color = "steelblue") +
  geom_text(aes(label = set), nudge_y = 0.02, size = 3) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "True value", y = "Estimated value")
```

## Shuffled-trial diagnostics

Shuffling trial order destroys the sequential Q-learning structure. If the model still "recovers" $\eta$ and $\tau$ from shuffled data, those parameters are not identified by temporal dependencies.

### Single-dataset shuffle test

Shuffle the original simulated data 200 times and refit.

```{r}
#| label: shuffled-fits
set.seed(2027)
n_shuffle <- 10

fit_shuffled_crl <- function(y, a) {
  idx <- sample.int(length(y))
  y_s <- y[idx]
  a_s <- a[idx]

  theta0 <- par_to_unconstrained_crl(init_pars)
  fit <- tryCatch(
    optim(par = theta0, fn = neg_loglik_crl, y = y_s, a = a_s,
          method = "Nelder-Mead",
          control = list(maxit = 10000, reltol = 1e-10)),
    error = \(e) list(par = rep(NA_real_, 4), convergence = 99)
  )

  est <- par_from_unconstrained_crl(fit$par)
  tibble(
    c = est$c, kappa = est$kappa,
    eta = est$eta, tau = est$tau,
    converged = fit$convergence == 0
  )
}

shuffle_fits <- with_cache("output/crl_shuffle_fits.rds", {
  purrr::map_dfr(seq_len(n_shuffle), \(i) fit_shuffled_crl(sim$y, sim$a))
})

shuffle_long <- shuffle_fits |>
  pivot_longer(cols = c(c, kappa, eta, tau),
               names_to = "parameter", values_to = "estimated")

ggplot(shuffle_long, aes(x = estimated)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(
    data = tibble(parameter = names(true_pars), true_value = unlist(true_pars)),
    aes(xintercept = true_value), color = "red", linetype = "dashed"
  ) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "Estimated value", y = "Count",
       caption = "Red dashed = generating parameter (unshuffled data)")
```

### Arc density: original vs shuffled-fit simulations

For each shuffled fit, simulate new data from the recovered parameters and compare the arc-width density to the original data.

```{r}
#| label: shuffled-arc-density
#| fig-width: 9
#| fig-height: 5
shuffled_sims <- purrr::pmap_dfr(
  shuffle_fits |> filter(converged),
  \(c, kappa, eta, tau, ...) {
    out <- run_arc_algorithm(
      algo_crl(eta = eta, tau = tau),
      c = c, kappa = kappa, init_arc = pi / 2, n_trials = 1000
    )
    tibble(arc = out$a)
  },
  .id = "run"
)

original_df <- tibble(arc = sim$a, source = "Original data")

ggplot() +
  geom_density(data = shuffled_sims, aes(x = arc, group = run),
               color = "steelblue", alpha = 0.3, linewidth = 0.4) +
  geom_density(data = original_df, aes(x = arc),
               color = "red", linewidth = 1.2) +
  labs(x = "Arc half-width (rad)", y = "Density",
       title = "Arc density: original (red) vs simulated from shuffled-fit parameters (blue)") +
  theme_minimal()
```

<!-- ### Shuffled recovery across parameter sets

Repeat the sanity-check recovery but with shuffled trials.

```{r}
#| label: shuffled-recovery
set.seed(2028)

fit_one_crl_shuffled <- function(c, kappa, eta, tau, n_trials = 1000) {
  sim <- run_arc_algorithm(
    arc_update_fun = algo_crl(eta = eta, tau = tau),
    c = c, kappa = kappa, init_arc = pi / 2, n_trials = n_trials
  )

  idx <- sample.int(length(sim$y))
  sim$y <- sim$y[idx]
  sim$a <- sim$a[idx]

  theta0 <- par_to_unconstrained_crl(list(
    c = 4, kappa = 4, eta = 0.1, tau = 0.1
  ))

  fit <- tryCatch(
    optim(par = theta0, fn = neg_loglik_crl, y = sim$y, a = sim$a,
          method = "Nelder-Mead",
          control = list(maxit = 10000, reltol = 1e-10)),
    error = \(e) list(par = rep(NA_real_, 4), convergence = 99)
  )

  est <- par_from_unconstrained_crl(fit$par)
  tibble(
    c_true = c, kappa_true = kappa, eta_true = eta, tau_true = tau,
    c_est = est$c, kappa_est = est$kappa, eta_est = est$eta, tau_est = est$tau,
    converged = fit$convergence == 0
  )
}

recovery_shuffled <- purrr::pmap_dfr(recovery_sets, fit_one_crl_shuffled)
```

```{r}
#| label: fig-shuffled-recovery
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Shuffled-trial recovery for CRL. Departure from the identity line for eta/tau indicates those parameters rely on temporal structure."
recovery_shuffled |>
  mutate(set = row_number()) |>
  pivot_longer(
    cols = -c(set, converged),
    names_to = c("parameter", ".value"),
    names_pattern = "(.+)_(true|est)"
  ) |>
  ggplot(aes(x = true, y = est)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_point(size = 3, color = "steelblue") +
  geom_text(aes(label = set), nudge_y = 0.02, size = 3) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "True value", y = "Estimated value")
``` -->
