---
title: "Model diagnostics: contraction-expansion (algo4) with logit noise, MLE"
format: html
---

```{r cache = FALSE}
#| label: setup
#| message: false
library(tidyverse)
library(bmm)
source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: {.callout-note}
## Prerequisites
This notebook fits the contraction-expansion model (algo4) with logit-scale noise, developed in [RL Algorithms: Exploration](reinforcement-learning-arc.qmd). The SDM model and optimality derivation are described in [Optimal Arc Width (SDM)](optimal-arc-response.qmd).
:::

## Task and model summary

On each trial a participant reports a remembered color (continuous response) and sets a symmetric **arc** of half-width $a_t \in (0,\pi)$ around it. Points are awarded as
$$
p_t = (\pi - a_t)\,\mathbf{1}\!\bigl[|y_t| \le a_t\bigr],
$$
where $y_t$ is the signed response error. Wider arcs are safer but earn fewer points, creating an accuracy–confidence tradeoff. [Empirically, participants' arc widths are near-optimal](optimal-arc-response.html) given their error distributions.

**Response errors** are modelled by the [Signal Discrimination Model](https://venpopov.github.io/bmm/articles/bmm_sdm_simple.html) (SDM):
$$
y_t \sim \text{SDM}(c,\kappa).
$$

**Deterministic arc-width update (algo4).** On each trial the arc contracts toward 0 on a hit and expands toward $\pi$ on a miss, with step-size $\delta \in (0,1)$:
$$
\tilde{a}_{t+1} = a_t\,(1-\delta) + \mathbf{1}\!\bigl[|y_t|>a_t\bigr]\,\delta\,\pi.
$$

**Logit-noise stochastic extension.** To make the model estimable, we add Gaussian noise on the unconstrained logit scale. Define
$$
\tilde{x}_{t+1} = \operatorname{logit}\!\left(\tilde{a}_{t+1}/\pi\right),
$$
then the stochastic transition is
$$
x_{t+1} = \tilde{x}_{t+1} + \varepsilon_t, \qquad \varepsilon_t \sim \mathcal{N}(0,\sigma_\varepsilon^2),
$$
so that $a_{t+1} = \pi\,\sigma(x_{t+1})$ is guaranteed to lie in $(0,\pi)$ without clipping.

**Joint log-likelihood.** Because $x_{t+1} \mid a_t, y_t$ is normal, the full log-likelihood decomposes as
$$
\ell(\theta) =
\underbrace{\sum_{t=1}^{T}\log f_{\text{SDM}}(y_t \mid c,\kappa)}_{\text{response errors}}
\;+\;
\underbrace{\sum_{t=1}^{T-1}\log\phi\!\left(x_{t+1};\; \tilde{x}_{t+1},\; \sigma_\varepsilon\right)}_{\text{arc transitions}},
$$
with fitted parameter vector $\theta=(c,\kappa,\delta,\sigma_\varepsilon)$. Below we verify identifiability via simulation-based parameter recovery and maximum-likelihood estimation with `optim`.

## Simulate data

Generate a single subject's worth of data from the algo4 logit-noise model with
known parameters. We use `run_arc_algorithm` which feeds SDM-generated response
errors `y` through the update function trial by trial.

```{r}
#| label: simulate-data
true_pars <- list(c = 6, kappa = 3, delta = 0.1, sigma_e = 0.2)

set.seed(4233)
sim <- run_arc_algorithm(
  arc_update_fun = algo4_logit_noise(
    delta = true_pars$delta,
    sigma_e = true_pars$sigma_e
  ),
  c = true_pars$c,
  kappa = true_pars$kappa,
  init_arc = pi / 8,
  n_trials = 1000
)

par(mfrow = c(1, 3))
plot(sim$a, type = "l", xlab = "Trial", ylab = "Arc width")
hist(sim$a, xlab = "Arc width")
hist(sim$y, xlab = "Response error")
par(mfrow = c(1, 1))
```

## Log-likelihood

The joint log-likelihood has two additive components:

1. **SDM component** — the density of each response error $y_t$
2. **Arc-transition component** — a normal density on the logit scale for the
   arc update from trial $t$ to $t+1$

The functions `algo4_loglik`, `par_to_unconstrained_algo4`,
`par_from_unconstrained_algo4`, and `neg_loglik_algo4` are defined in
`R/arc-models.R` (loaded above). Parameters $c$, $\kappa$, and
$\sigma_\varepsilon$ are strictly positive (estimated on the log scale); $\delta
\in (0,1)$ is estimated on the logit scale.

## Fit with `optim`

We initialise at perturbed values (not the truth) to verify that the optimiser
can recover the generating parameters.

```{r}
#| label: fit-optim
init_pars <- list(c = 3, kappa = 5, delta = 0.05, sigma_e = 0.5)
theta0 <- par_to_unconstrained_algo4(init_pars)

fit <- optim(
  par = theta0,
  fn = neg_loglik_algo4,
  y = sim$y,
  a = sim$a,
  method = "Nelder-Mead",
  control = list(maxit = 5000, reltol = 1e-10)
)

fit$convergence  # 0 = success
```

## Compare estimated vs true parameters

```{r}
#| label: compare-params
estimated <- par_from_unconstrained_algo4(fit$par)

comparison <- tibble(
  parameter = names(true_pars),
  initial = unlist(init_pars),
  true_value = unlist(true_pars),
  estimated = unlist(estimated)
)
comparison
```

## Likelihood profile at the MLE

A quick sanity check: evaluate the likelihood over a grid around each MLE
estimate while holding the others fixed, to confirm we are at a peak.

```{r}
#| label: likelihood-profile
#| fig-width: 9
#| fig-height: 6
best_theta <- fit$par
par_names <- c("c", "kappa", "delta", "sigma_e")

profiles <- purrr::map_dfr(seq_along(par_names), \(j) {
  grid <- seq(best_theta[j] - 1, best_theta[j] + 1, length.out = 80)
  ll_vals <- sapply(grid, \(v) {
    theta_tmp <- best_theta
    theta_tmp[j] <- v
    -neg_loglik_algo4(theta_tmp, y = sim$y, a = sim$a)
  })

  if (par_names[j] == "delta") {
    value <- inv_logit(grid)
  } else {
    value <- exp(grid)
  }

  tibble(
    parameter = par_names[j],
    theta_unconstrained = grid,
    value = value,
    loglik = ll_vals
  )
})

ggplot(profiles, aes(x = value, y = loglik)) +
  geom_line() +
  geom_vline(data = comparison, aes(xintercept = estimated), linetype = "dashed", color = "blue") +
  geom_vline(data = comparison, aes(xintercept = true_value), linetype = "dotted", color = "red") +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "Parameter value", y = "Log-likelihood",
       caption = "Blue dashed = MLE, Red dotted = true value")
```

## Parameter recovery

Sample 10000 parameter sets uniformly, simulate data for each, fit via MLE, and
check whether the true values are recovered.

```{r}
#| label: param-recovery
n_recovery <- 10000
n_trials <- 1000
set.seed(2026)

recovery_true <- tibble(
  iter = seq_len(n_recovery),
  c = runif(n_recovery, 1, 10),
  kappa = runif(n_recovery, 1, 10),
  delta = runif(n_recovery, 0.01, 0.5),
  sigma_e = runif(n_recovery, 0.01, 1)
)

fit_one <- function(iter, c, kappa, delta, sigma_e) {
  sim <- run_arc_algorithm(
    arc_update_fun = algo4_logit_noise(delta = delta, sigma_e = sigma_e),
    c = c, kappa = kappa, init_arc = pi / 2, n_trials = n_trials
  )

  theta0 <- par_to_unconstrained_algo4(list(
    c = 4, kappa = 4, delta = 0.1, sigma_e = 0.3
  ))

  fit <- tryCatch(
    optim(par = theta0, fn = neg_loglik_algo4, y = sim$y, a = sim$a,
          method = "Nelder-Mead",
          control = list(maxit = 5000, reltol = 1e-10)),
    error = \(e) list(par = rep(NA_real_, 4), convergence = 99)
  )

  est <- par_from_unconstrained_algo4(fit$par)
  tibble(
    iter = iter,
    c_est = est$c, kappa_est = est$kappa,
    delta_est = est$delta, sigma_e_est = est$sigma_e,
    converged = fit$convergence == 0
  )
}

recovery_est <- with_cache("output/algo4_recovery_est.rds", {
  purrr::pmap_dfr(recovery_true, fit_one)
})
```

Tag iterations where at least one estimate exceeds 4× the max true generating value:

```{r}
#| label: tag-recovery
est_cols <- c("c_est", "kappa_est", "delta_est", "sigma_e_est")
true_cols <- c("c", "kappa", "delta", "sigma_e")
thresholds <- 4 * sapply(true_cols, \(col) max(recovery_true[[col]]))
outlier_mat <- sapply(seq_along(est_cols), \(j) {
  recovery_est[[est_cols[j]]] > thresholds[j]
})
recovery_est$bad <- rowSums(outlier_mat, na.rm = TRUE) > 0
```

Join true and estimated values and reshape for plotting:

```{r}
#| label: build-recovery-df
recovery_df <- recovery_true |>
  left_join(recovery_est, by = "iter") |>
  pivot_longer(
    cols = c(c, kappa, delta, sigma_e),
    names_to = "parameter",
    values_to = "true_value"
  ) |>
  mutate(
    estimated = case_when(
      parameter == "c" ~ c_est,
      parameter == "kappa" ~ kappa_est,
      parameter == "delta" ~ delta_est,
      parameter == "sigma_e" ~ sigma_e_est
    )
  ) |>
  select(iter, parameter, true_value, estimated, converged, bad)

cat("Converged:", sum(recovery_est$converged), "/", n_recovery,
    sprintf("(%.1f%%)", 100 * mean(recovery_est$converged)), "\n")
cat("Bad parameter estimates:", sum(recovery_est$bad), "/", n_recovery,
    sprintf("(%.1f%%)", 100 * mean(recovery_est$bad)), "\n")
```

Correlations between true and estimated values (converged runs only):

```{r}
#| label: recovery-correlations
recovery_df |>
  filter(converged, !bad) |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated),
    rmse = sqrt(mean((true_value - estimated)^2)),
    bias = mean(estimated - true_value),
    .groups = "drop"
  )
```

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery
#| fig-cap: "Parameter recovery for the contraction-expansion model with logit noise. Each point is one simulated dataset. The dashed line is the identity."

plot_df <- recovery_df |> filter(converged, !bad)

cor_labels <- plot_df |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated),
    x = -Inf, y = Inf,
    .groups = "drop"
  ) |>
  mutate(label = sprintf("r = %.2f", r))

plot_df |>
  ggplot(aes(x = true_value, y = estimated)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_point(aes(color = converged), alpha = 0.1, size = 1.5) +
  geom_text(data = cor_labels, aes(x = x, y = y, label = label),
            hjust = -0.1, vjust = 1.3, inherit.aes = FALSE, size = 4) +
  facet_wrap(~parameter, scales = "free") +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "red")) +
  labs(x = "True value", y = "Estimated value", color = "Converged")
```

Binned summary: for each parameter, true values are split into 20 equal-width
bins. Within each bin, we show the median estimate (line) and 80%/95% intervals
(ribbons).

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery-binned
#| fig-cap: "Binned parameter recovery. Line = median estimate per bin; dark ribbon = 80% interval; light ribbon = 95% interval."

binned_df <- plot_df |>
  group_by(parameter) |>
  mutate(bin = cut(true_value, breaks = 20, labels = FALSE)) |>
  group_by(parameter, bin) |>
  summarise(
    true_mid = mean(true_value),
    median_est = median(estimated),
    q025 = quantile(estimated, 0.025),
    q10 = quantile(estimated, 0.10),
    q90 = quantile(estimated, 0.90),
    q975 = quantile(estimated, 0.975),
    .groups = "drop"
  )

ggplot(binned_df, aes(x = true_mid)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = q10, ymax = q90), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = median_est), color = "steelblue", linewidth = 0.8) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "True value", y = "Estimated value")
```

## Shuffled-trial diagnostics

This test checks whether the model can recover learning-related parameters
from variability alone. Shuffling the trials should destroy the sequential
dependencies that the algorithm relies on, so any recovery here reflects
information not tied to trial-to-trial structure.

### Shuffled-trial fits

Shuffle the trial order of the simulated dataset 1000 times and fit the model
each time. This destroys trial-to-trial dependencies while keeping the same
observed marginal distributions.

```{r}
#| label: shuffled-trial-fits
set.seed(2027)
n_shuffle <- 1000

fit_shuffled_once <- function(y, a) {
  idx <- sample.int(length(y))
  y_s <- y[idx]
  a_s <- a[idx]

  theta0 <- par_to_unconstrained_algo4(init_pars)
  fit <- tryCatch(
    optim(par = theta0, fn = neg_loglik_algo4, y = y_s, a = a_s,
          method = "Nelder-Mead",
          control = list(maxit = 5000, reltol = 1e-10)),
    error = \(e) list(par = rep(NA_real_, 4), convergence = 99)
  )

  est <- par_from_unconstrained_algo4(fit$par)
  tibble(
    c = est$c, kappa = est$kappa,
    delta = est$delta, sigma_e = est$sigma_e,
    converged = fit$convergence == 0
  )
}

shuffle_fits <- with_cache("output/algo4_shuffle_fits.rds", {
  purrr::map_dfr(seq_len(n_shuffle), \(i) fit_shuffled_once(sim$y, sim$a))
})

shuffle_long <- shuffle_fits |>
  pivot_longer(
    cols = c(c, kappa, delta, sigma_e),
    names_to = "parameter",
    values_to = "estimated"
  )

ggplot(shuffle_long, aes(x = estimated)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(
    data = tibble(parameter = names(true_pars), true_value = unlist(true_pars)),
    aes(xintercept = true_value), color = "red", linetype = "dashed"
  ) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "Estimated value", y = "Count",
       caption = "Red dashed = generating parameter (unshuffled data)")
```

### Parameter recovery with shuffled trials

Repeat the recovery analysis, but shuffle the trial order once before fitting
each simulated dataset.

```{r}
#| label: shuffled-recovery-fit
fit_one_shuffled <- function(iter, c, kappa, delta, sigma_e) {
  sim <- run_arc_algorithm(
    arc_update_fun = algo4_logit_noise(delta = delta, sigma_e = sigma_e),
    c = c, kappa = kappa, init_arc = pi / 2, n_trials = n_trials
  )

  idx <- sample.int(length(sim$y))
  sim$y <- sim$y[idx]
  sim$a <- sim$a[idx]

  theta0 <- par_to_unconstrained_algo4(list(
    c = 4, kappa = 4, delta = 0.1, sigma_e = 0.3
  ))

  fit <- tryCatch(
    optim(par = theta0, fn = neg_loglik_algo4, y = sim$y, a = sim$a,
          method = "Nelder-Mead",
          control = list(maxit = 5000, reltol = 1e-10)),
    error = \(e) list(par = rep(NA_real_, 4), convergence = 99)
  )

  est <- par_from_unconstrained_algo4(fit$par)
  tibble(
    iter = iter,
    c_est = est$c, kappa_est = est$kappa,
    delta_est = est$delta, sigma_e_est = est$sigma_e,
    converged = fit$convergence == 0
  )
}

recovery_est_shuffled <- with_cache("output/algo4_recovery_est_shuffled.rds", {
  purrr::pmap_dfr(recovery_true, fit_one_shuffled)
})

outlier_mat_shuffled <- sapply(seq_along(est_cols), \(j) {
  recovery_est_shuffled[[est_cols[j]]] > thresholds[j]
})
recovery_est_shuffled$bad <- rowSums(outlier_mat_shuffled, na.rm = TRUE) > 0

recovery_df_shuffled <- recovery_true |>
  left_join(recovery_est_shuffled, by = "iter") |>
  pivot_longer(
    cols = c(c, kappa, delta, sigma_e),
    names_to = "parameter",
    values_to = "true_value"
  ) |>
  mutate(
    estimated = case_when(
      parameter == "c" ~ c_est,
      parameter == "kappa" ~ kappa_est,
      parameter == "delta" ~ delta_est,
      parameter == "sigma_e" ~ sigma_e_est
    )
  ) |>
  select(iter, parameter, true_value, estimated, converged, bad)

cat("Converged:", sum(recovery_est_shuffled$converged), "/", n_recovery,
    sprintf("(%.1f%%)", 100 * mean(recovery_est_shuffled$converged)), "\n")
cat("Bad parameter estimates:", sum(recovery_est_shuffled$bad), "/", n_recovery,
    sprintf("(%.1f%%)", 100 * mean(recovery_est_shuffled$bad)), "\n")
```

```{r}
#| label: shuffled-recovery-correlations
recovery_df_shuffled |>
  filter(converged, !bad) |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated),
    rmse = sqrt(mean((true_value - estimated)^2)),
    bias = mean(estimated - true_value),
    .groups = "drop"
  )
```

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery-shuffled
#| fig-cap: "Parameter recovery with shuffled trials. Each point is one simulated dataset. The dashed line is the identity."

plot_df_shuffled <- recovery_df_shuffled |> filter(converged, !bad)

cor_labels_shuffled <- plot_df_shuffled |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated),
    x = -Inf, y = Inf,
    .groups = "drop"
  ) |>
  mutate(label = sprintf("r = %.2f", r))

plot_df_shuffled |>
  ggplot(aes(x = true_value, y = estimated)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_point(aes(color = converged), alpha = 0.1, size = 1.5) +
  geom_text(data = cor_labels_shuffled, aes(x = x, y = y, label = label),
            hjust = -0.1, vjust = 1.3, inherit.aes = FALSE, size = 4) +
  facet_wrap(~parameter, scales = "free") +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "red")) +
  labs(x = "True value", y = "Estimated value", color = "Converged")
```

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery-binned-shuffled
#| fig-cap: "Binned parameter recovery with shuffled trials. Line = median estimate per bin; dark ribbon = 80% interval; light ribbon = 95% interval."

binned_df_shuffled <- plot_df_shuffled |>
  group_by(parameter) |>
  mutate(bin = cut(true_value, breaks = 20, labels = FALSE)) |>
  group_by(parameter, bin) |>
  summarise(
    true_mid = mean(true_value),
    median_est = median(estimated),
    q025 = quantile(estimated, 0.025),
    q10 = quantile(estimated, 0.10),
    q90 = quantile(estimated, 0.90),
    q975 = quantile(estimated, 0.975),
    .groups = "drop"
  )

ggplot(binned_df_shuffled, aes(x = true_mid)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = q10, ymax = q90), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = median_est), color = "steelblue", linewidth = 0.8) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "True value", y = "Estimated value")
```
