---
title: "Exp 1: By-Subject Diagnostics"
---

This notebook shows per-subject distributions of point response error and arc width from Experiment 1. The goal is to visually inspect individual differences — some participants may have much tighter or wider distributions than others, or may show unusual patterns (e.g., near-uniform responses suggesting inattention). There are 20 participants, some contributing to more than one experimental session.

::: {.callout-note}
## Data source
Preprocessed data: `output/exp1_data.csv`.
:::

```{r}
#| message: false
library(tidyverse)
exp1 <- read.csv("output/exp1_data.csv")
```

## Point response error

```{r}
#| fig-width: 10
#| fig-height: 10
exp1 |> 
  mutate(subj_exp = glue::glue("subj_{subject}_exp_{experimentorder}")) |> 
  ggplot(aes(resperr)) + 
  geom_histogram(bins = 60) +
  facet_wrap(~ subj_exp) +
  theme_classic()
```

## Arc widths

```{r}
#| fig-width: 10
#| fig-height: 10
exp1 |> 
  mutate(subj_exp = glue::glue("subj_{subject}_exp_{experimentorder}")) |> 
  ggplot(aes(arc)) + 
  geom_histogram(bins = 60) +
  facet_wrap(~ subj_exp) +
  theme_classic()
```

## Side by side

```{r}
#| fig-width: 4
#| fig-height: 80
exp1 |> 
  mutate(subj_exp = glue::glue("S{subject}_E{experimentorder}")) |> 
  gather(measure, value, arc, resperr) |> 
  ggplot(aes(value)) + 
  geom_histogram(bins = 60) +
  facet_grid(subj_exp ~ measure, scales = "free_x") +
  theme_bw()
```
