---
title: "Fitting CRL to experimental data (MLE)"
format: html
---

```{r cache = FALSE}
#| label: init
#| message: false
library(tidyverse)
library(patchwork)
library(bmm)
source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: {.callout-note}
## Prerequisites
This notebook fits the Counterfactual Q-Learning model (CRL) to real data via
MLE. The model is developed in [Counterfactual RL](counterfactual-rl.qmd);
simulation-based diagnostics are in [CRL MLE diagnostics](diagnostics-crl-mle.qmd).
:::

## Load data

```{r}
#| label: load-data
exp1 <- read_csv("output/exp1_data.csv", show_col_types = FALSE)
dat <- exp1 |>
  filter(exp_type == "Time", experimentorder == 1) |>
  select(subject, blocknumber, trialnumber, setsize, resperr, arc, points) |>
  arrange(subject, blocknumber, trialnumber) |>
  mutate(half_arc = arc / 2, half_arc = if_else(half_arc > 0, half_arc, 1e-6))

subject_ids <- sort(unique(dat$subject))
subject_data <- split(dat, dat$subject)
```

## Fit individual participants

```{r}
#| label: fit-one-subject
fit_crl_mle <- function(data, n_restarts = 3) {
  y <- data$resperr
  a <- data$half_arc

  best <- list(value = Inf)

  for (r in seq_len(n_restarts)) {
    theta0 <- par_to_unconstrained_crl(list(
      c = runif(1, 2, 8),
      kappa = runif(1, 1, 6),
      eta = runif(1, 0.02, 0.3),
      tau = runif(1, 0.02, 0.2)
    ))

    fit <- tryCatch(
      optim(par = theta0, fn = neg_loglik_crl, y = y, a = a,
            method = "Nelder-Mead",
            control = list(maxit = 15000, reltol = 1e-10)),
      error = \(e) list(par = rep(NA_real_, 4), value = Inf, convergence = 99)
    )

    if (fit$value < best$value) best <- fit
  }

  est <- par_from_unconstrained_crl(best$par)
  tibble(
    c = est$c, kappa = est$kappa, eta = est$eta, tau = est$tau,
    loglik = -best$value, converged = best$convergence == 0
  )
}
```

```{r}
#| label: fit-all-subjects
set.seed(2026)

fits <- with_cache("output/crl_mle_fits.rds", {
  imap_dfr(subject_data, \(data, sid) {
    fit_crl_mle(data) |> mutate(subject = sid, .before = 1)
  })
})

fits
```

## Parameter estimates across participants

```{r}
#| label: fig-param-estimates
#| fig-width: 10
#| fig-height: 5
fits |>
  pivot_longer(cols = c(c, kappa, eta, tau),
               names_to = "parameter", values_to = "estimate") |>
  ggplot(aes(x = subject, y = estimate)) +
  geom_point(size = 2, color = "steelblue") +
  facet_wrap(~parameter, scales = "free_y") +
  labs(x = "Subject", y = "MLE estimate") +
  theme_minimal()
```

## Posterior predictive checks

Simulate data from each participant's fitted parameters and compare arc-width
and response-error distributions.

```{r}
#| label: ppc-helpers
crl_sim_from_pars <- function(pars, n_trials, init_arc = pi / 2) {
  run_arc_algorithm(
    arc_update_fun = algo_crl(eta = pars$eta, tau = pars$tau),
    c = pars$c, kappa = pars$kappa,
    init_arc = init_arc, n_trials = n_trials
  )
}

plot_crl_ppc <- function(pars, data, n_sims = 15, title = "") {
  sims <- replicate(n_sims,
    crl_sim_from_pars(pars, n_trials = nrow(data), init_arc = data$half_arc[1]),
    simplify = FALSE
  )
  sim_col <- adjustcolor("red", alpha.f = 0.3)

  optim_arc <- find_optimum_sdm(pars$c, pars$kappa)$alpha
  obs_median <- median(data$half_arc)
  pred_medians <- vapply(sims, \(s) median(s$a), numeric(1))

  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar), add = TRUE)
  par(mfrow = c(1, 2))

  plot(density(data$half_arc), main = paste0("Arc width – ", title))
  for (s in sims) lines(density(s$a), col = sim_col)
  lines(density(data$half_arc), lwd = 2)
  abline(v = optim_arc, col = "blue", lwd = 2, lty = 2)
  abline(v = obs_median, col = "black", lwd = 2, lty = 3)
  abline(v = median(pred_medians), col = "red", lwd = 2, lty = 3)
  legend("topright",
    c("Data", "Simulated", "Optimal", "Obs med", "Pred med"),
    col = c("black", "red", "blue", "black", "red"),
    lty = c(1, 1, 2, 3, 3), lwd = c(1, 1, 2, 2, 2), bty = "n", cex = 0.8)

  plot(density(data$resperr), main = paste0("Response error – ", title))
  for (s in sims) lines(density(s$y), col = sim_col)
  lines(density(data$resperr), lwd = 2)
  legend("topright", c("Data", "Simulated"), col = c("black", "red"), lty = 1, bty = "n")
}
```

```{r}
#| label: ppc-all
#| fig-width: 10
#| fig-height: 4
#| fig-show: hold
iwalk(subject_data, \(data, sid) {
  pars <- fits |> filter(subject == sid)
  plot_crl_ppc(pars, data, n_sims = 15, title = paste("Subject", sid))
})
```

## Predicted vs observed arc summary statistics

```{r}
#| label: ppc-summary-stats
#| fig-width: 10
#| fig-height: 5
ppc_summary <- imap_dfr(subject_data, \(data, sid) {
  pars <- fits |> filter(subject == sid)
  sims <- replicate(50,
    crl_sim_from_pars(pars, n_trials = nrow(data), init_arc = data$half_arc[1]),
    simplify = FALSE
  )
  tibble(
    subject = sid,
    obs_median = median(data$half_arc),
    obs_mad = mad(data$half_arc),
    pred_median = map_dbl(sims, \(s) median(s$a)),
    pred_mad = map_dbl(sims, \(s) mad(s$a))
  )
})

ppc_smry <- ppc_summary |>
  group_by(subject, obs_median, obs_mad) |>
  summarise(
    pred_median_mid = median(pred_median),
    pred_median_lo = quantile(pred_median, 0.025),
    pred_median_hi = quantile(pred_median, 0.975),
    pred_mad_mid = median(pred_mad),
    pred_mad_lo = quantile(pred_mad, 0.025),
    pred_mad_hi = quantile(pred_mad, 0.975),
    .groups = "drop"
  )

p1 <- ggplot(ppc_smry, aes(x = obs_median, y = pred_median_mid)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = pred_median_lo, ymax = pred_median_hi)) +
  labs(x = "Observed median arc", y = "Predicted median arc",
       title = "Median arc width") +
  theme_minimal()

p2 <- ggplot(ppc_smry, aes(x = obs_mad, y = pred_mad_mid)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = pred_mad_lo, ymax = pred_mad_hi)) +
  labs(x = "Observed MAD arc", y = "Predicted MAD arc",
       title = "MAD of arc width") +
  theme_minimal()

p1 + p2
```
