---
title: "Fitting algo4 to experimental data (Bayesian)"
format: html
execute: 
  enabled: false
---

```{r cache = FALSE}
#| label: setup
#| message: false
library(tidyverse)
library(brms)
library(bmm)
source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: {.callout-note}
## Prerequisites
This notebook fits the contraction-expansion model (algo4) with logit-scale
noise to real experimental data using full Bayesian inference. Model diagnostics
and parameter recovery are in
[algo4 Bayesian diagnostics](diagnostics-algo4-bayes.qmd). The MLE counterpart
is in [algo4 MLE diagnostics](diagnostics-algo4-mle.qmd).
:::

## Model setup

```{r}
#| label: define-family-stanvars-priors
sdm_algo4 <- custom_family(
  "sdm_algo4",
  dpars = c("mu", "c", "kappa", "delta", "asigma"),
  links = c("tan_half", "identity", "log", "logit", "log"),
  type = "real",
  loop = FALSE,
  vars = "vreal1"
)

stanvars <- sdm_base_stanvars() +
  stanvar(
    scode = read_file("stan/sdm_algo4_likelihood.stan"),
    block = "functions"
  )

prior <- set_prior("constant(0)", dpar = "mu", class = "Intercept") +
  set_prior("student_t(5, 2, 0.75)", dpar = "c", class = "Intercept") +
  set_prior("student_t(5, 1.75, 0.75)", dpar = "kappa", class = "Intercept") +
  set_prior("normal(-3.5, 2)", dpar = "delta", class = "Intercept") +
  set_prior("normal(-3.5, 2)", dpar = "asigma", class = "Intercept")
```

## Load data

```{r}
#| label: load-data
exp1 <- read_csv("output/exp1_data.csv", show_col_types = FALSE)
dat <- exp1 |>
  filter(exp_type == "Time", experimentorder == 1) |>
  select(subject, experimentorder, blocknumber, trialnumber, setsize, resperr, arc, points) |>
  mutate(half_arc = arc / 2)
```

## Single-subject fit

Select one subject and remove trials with zero arc width:

```{r}
#| label: select-subject
s1 <- dat |> filter(subject == 6, arc > 0)
```

```{r}
#| label: fit-subject
fit_s1 <- brm(
  bf(resperr | vreal(half_arc) ~ 1, c ~ 1, kappa ~ 1, delta ~ 1, asigma ~ 1),
  prior = prior,
  data = s1, family = sdm_algo4, stanvars = stanvars,
  backend = "cmdstanr", cores = 4, init = 0,
  file = "output/fit-algo4-bayes-s6"
)
```

```{r}
#| label: fit-summary
summary(fit_s1)
```

```{r}
#| label: fit-plot
#| fig-width: 10
#| fig-height: 6
plot(fit_s1)
```

## Posterior predictive check

```{r}
#| label: extract-fit-params
sdm4_pars_from_fit <- function(fit) {
  est <- fixef(fit)[-1, 1]
  list(
    c = exp(est["c_Intercept"]),
    kappa = exp(est["kappa_Intercept"]),
    delta = inv_logit(est["delta_Intercept"]),
    sigma_e = exp(est["asigma_Intercept"])
  )
}
```

```{r}
#| label: pp-arc-plot
#| fig-width: 10
#| fig-height: 8
pars <- sdm4_pars_from_fit(fit_s1)

sim_s1 <- run_arc_algorithm(
  arc_update_fun = algo4_logit_noise(
    delta = pars$delta,
    sigma_e = pars$sigma_e
  ),
  c = pars$c,
  kappa = pars$kappa,
  init_arc = s1$half_arc[1],
  n_trials = nrow(s1)
)

par(mfrow = c(2, 2))
plot(s1$half_arc, type = "l", xlab = "Trial", ylab = "Arc half-width",
     main = "Arc trajectory", ylim = c(0, pi))
lines(sim_s1$a, col = "red")
legend("topright", c("Data", "Posterior"), col = c("black", "red"), lty = 1)

plot(density(s1$half_arc), main = "Arc width distribution")
lines(density(sim_s1$a), col = "red")
abline(v = median(s1$half_arc), lty = 2)
abline(v = median(sim_s1$a), col = "red", lty = 2)
legend("topright", c("Data", "Posterior"), col = c("black", "red"), lty = 1)

plot(density(s1$resperr), main = "Response error distribution")
lines(density(sim_s1$y), col = "red")
legend("topright", c("Data", "Posterior"), col = c("black", "red"), lty = 1)
par(mfrow = c(1, 1))
```
