---
title: "Fitting algo4 to experimental data (Bayesian)"
format: html
---

```{r cache = FALSE}
#| label: init
#| message: false
library(tidyverse)
library(ggdist)
library(brms)
library(bmm)
source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

::: {.callout-note}
## Prerequisites
This notebook fits the contraction-expansion model (algo4) with logit-scale
noise to real experimental data using full Bayesian inference. Model diagnostics
and parameter recovery are in
[algo4 Bayesian diagnostics](diagnostics-algo4-bayes.qmd). The MLE counterpart
is in [algo4 MLE diagnostics](diagnostics-algo4-mle.qmd).
:::

## Model setup

```{r}
#| label: define-family-stanvars-priors
sdm_algo4 <- custom_family(
  "sdm_algo4",
  dpars = c("mu", "c", "kappa", "delta", "asigma"),
  links = c("tan_half", "identity", "log", "logit", "log"),
  type = "real",
  loop = FALSE,
  vars = c("vreal1", "vreal2")
)

stanvars <- sdm_base_stanvars() +
  stanvar(
    scode = read_file("stan/sdm_algo4_likelihood.stan"),
    block = "functions"
  )

prior <- set_prior("constant(0)", dpar = "mu", class = "Intercept") +
  set_prior("student_t(5, 2, 0.75)", dpar = "c", class = "Intercept") +
  set_prior("student_t(5, 1.75, 0.75)", dpar = "kappa", class = "Intercept") +
  set_prior("normal(-3.5, 2)", dpar = "delta", class = "Intercept") +
  set_prior("normal(-3.5, 2)", dpar = "asigma", class = "Intercept")


fit_algo4_bayes <- function(data, file) {
  brm(
    bf(resperr | vreal(half_arc, arc_next) ~ 1, c ~ 1, kappa ~ 1, delta ~ 1, asigma ~ 1),
    prior = prior,
    data = data, family = sdm_algo4, stanvars = stanvars,
    backend = "cmdstanr", cores = 4, init = 0, file = file,
    refresh = 0
  )
}
```

## Load data

```{r}
#| label: load-data
exp1 <- read_csv("output/exp1_data.csv", show_col_types = FALSE)
dat <- exp1 |>
  filter(exp_type == "Time", experimentorder == 1) |>
  select(subject, experimentorder, blocknumber, trialnumber, setsize, resperr, arc, points) |>
  arrange(subject, blocknumber, trialnumber) |>
  mutate(half_arc = arc / 2, half_arc = if_else(half_arc > 0, half_arc, 1e-6)) |>
  group_by(subject) |>
  mutate(arc_next = lead(half_arc, default = -1)) |>
  ungroup()
```

## Single-subject fit (sanity check)

Select one subject to test the pipeline:

```{r}
#| label: fit-subject
s1 <- dat |> filter(subject == 6)
fit_s1 <- fit_algo4_bayes(s1, file = "output/fit-algo4-bayes-s6")
```

```{r}
#| label: fit-summary
summary(fit_s1)
```

## Plot posteriors on native scale

```{r}
#| warning: false
#| label: native-posterior-plot
algo4_posterior_draws <- function(fit, n_draws = NULL) {
  draws <- as_draws_df(fit) |> 
    select(starts_with("Intercept"), -Intercept)

  colnames(draws) <- gsub("Intercept_", "", colnames(draws))

  if (is.null(n_draws)) {
    idx <- seq_len(nrow(draws))
  } else {
    idx <- sample(seq_len(nrow(draws)), size = n_draws)
  }

  draws |>
    mutate(c = exp(c), kappa = exp(kappa), delta = inv_logit(delta), asigma = exp(asigma)) |> 
    slice(idx)
}

algo4_posterior_draws(fit_s1) |> 
  pivot_longer(cols = everything(), names_to = "parameter", values_to = "est")  |> 
  ggplot(aes(est)) +
  stat_halfeye(normalize = "groups") +
  facet_wrap(~parameter, scales = "free")
```

## Posterior predictive check

```{r}
#| label: ppc-helpers
#| code-fold: true
algo4_sim_from_draws <- function(draws, data) {
  purrr::pmap(
    draws,
    \(c, kappa, delta, asigma) {
      run_arc_algorithm(
        arc_update_fun = algo4_logit_noise(delta = delta, sigma_e = asigma),
        c = c,
        kappa = kappa,
        init_arc = data$half_arc[1],
        n_trials = nrow(data)
      )
    }
  )
}

plot_algo4_ppc <- function(fit, data, n_draws = 15) {
  draws <- algo4_posterior_draws(fit, n_draws = n_draws)
  sims <- algo4_sim_from_draws(draws, data)
  posterior_col <- grDevices::adjustcolor("red", alpha.f = 0.3)

  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar), add = TRUE)
  par(mfrow = c(1, 2))

  plot(density(data$half_arc), main = "Arc width distribution")
  for (sim in sims) {
    lines(density(sim$a), col = posterior_col)
  }
  lines(density(data$half_arc), lwd = 2)
  legend("topright", c("Data", "Posterior draws"), col = c("black", "red"), lty = 1, bty = "n")

  plot(density(data$resperr), main = "Response error distribution")
  for (sim in sims) {
    lines(density(sim$y), col = posterior_col)
  }
  lines(density(data$resperr), lwd = 2)
  legend("topright", c("Data", "Posterior draws"), col = c("black", "red"), lty = 1, bty = "n")

  invisible(sims)
}

plot_algo4_ppc_trajectories <- function(fit, data, n_draws = 15) {
  draws <- algo4_posterior_draws(fit, n_draws = n_draws)
  sims <- algo4_sim_from_draws(draws, data)

  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar), add = TRUE)
  par(mfrow = c(4, 4))

  plot(data$half_arc, type = "l", xlab = "Trial", ylab = "Arc half-width",
       main = "Data", ylim = c(0, pi), col = "blue")

  for (sim in sims) {
    plot(sim$a, type = "l", xlab = "Trial", ylab = "Arc half-width",
         main = "Posterior", ylim = c(0, pi))
  }

  invisible(sims)
}

summarize_algo4_posterior <- function(fit, subject_id) {
  algo4_posterior_draws(fit) |>
    pivot_longer(cols = everything(), names_to = "parameter", values_to = "est") |>
    group_by(parameter) |>
    summarise(
      median = median(est),
      lower = quantile(est, 0.025),
      upper = quantile(est, 0.975),
      .groups = "drop"
    ) |>
    mutate(subject = subject_id, .before = 1)
}
```

We plot 15 posterior predictive distributions for response error and half arc-width. While the response distribution fits with the posterior prediction, the arc distribution from the model has too much variance.

```{r}
#| label: pp-arc-plot
#| fig-width: 10
#| fig-height: 4
plot_algo4_ppc(fit_s1, s1, n_draws = 15)
```

Here are different predicted arc trajectories with parameters drawn from the posterior vs the data (top left panel):

```{r}
#| label: pp-arc-traj-grid
#| fig-width: 10
#| fig-height: 10
plot_algo4_ppc_trajectories(fit_s1, s1, n_draws = 15)
```


## Unpooled fits to all subjects individually

```{r}
#| label: fit-all-subjects
#| output: false
subject_ids <- as.character(sort(unique(dat$subject)))

subject_data <- subject_ids |>
  set_names() |>
  map(\(subject_id) dat |> filter(subject == subject_id))

fits_all <- imap(
  subject_data,
  \(data, subject_id) fit_algo4_bayes(data, file = paste0("output/fit-algo4-bayes-s", subject_id))
)
```

Posterior 95 BCI for all subjects and parameters

```{r}
#| label: posterior-summary
#| warning: false
posterior_smry <- imap_dfr(
  fits_all,
  \(fit, subject_id) summarize_algo4_posterior(fit, subject_id)
)

ggplot(posterior_smry, aes(subject, y = median)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper)) +
  facet_wrap(~parameter, scales = "free")
```

Posterior predictives for all subjects shows bad fit to the arc distribution

```{r}
#| label: ppc-all-subjects
#| warning: false
#| fig-width: 10
#| fig-height: 4
#| fig-show: hold
iwalk(
  fits_all,
  \(fit, subject_id) plot_algo4_ppc(fit, subject_data[[as.character(subject_id)]], n_draws = 15)
)
```

## Hierarchical fit

The likelihood now handles subject boundaries via the `arc_next` column: the last
trial of each subject has `arc_next = -1`, so the Stan code skips the arc-transition
term for those rows. This makes it safe to stack all subjects' data.

Skip for now, will run later

```{r}
#| label: fit-hierarchical
hfit <- brm(
  bf(
    resperr | vreal(half_arc, arc_next) ~ 1,
    c ~ 1 + (1 | subject),
    kappa ~ 1 + (1 | subject),
    delta ~ 1 + (1 | subject),
    asigma ~ 1 + (1 | subject)
  ),
  prior = prior,
  data = dat, family = sdm_algo4, stanvars = stanvars,
  backend = "cmdstanr", cores = 4, init = 0, refresh = 0,
  file = "output/fit-algo4-bayes-hierarchical"
)
summary(hfit)
```
