---
title: "Model fits: noisy-gradient (algo5), MLE"
format: html
---

```{r setup}
#| message: false
library(tidyverse)
library(bmm)
source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

## Task and model summary

On each trial a participant reports a remembered color (continuous response) and sets a symmetric **arc** of half-width $a_t \in (0,\pi)$ around it. Points are awarded as
$$
p_t = (\pi - a_t)\,\mathbf{1}\!\bigl[|y_t| \le a_t\bigr],
$$
where $y_t$ is the signed response error. Wider arcs are safer but earn fewer points, creating an accuracy–confidence tradeoff. [Empirically, participants' arc widths are near-optimal](optimal-arc-response.html) given their error distributions.

**Response errors** are modelled by the Signal Detection Model (SDM):
$$
y_t \sim \text{SDM}(c,\kappa).
$$

**Arc-width updates** follow a noisy stochastic-gradient rule on the logit-reparameterized arc $x_t = \operatorname{logit}(a_t/\pi)$. We do this on the unconstrained $x_t$ so that despite adding random noise, responses that get mapped back to $a_{t+1} = \operatorname{invlogit}(x_{t+1})\pi$ are in the right range, $(0,\pi)$, without needing to truncate.  Define a soft hit indicator and gradient:
$$
s_t = \sigma\!\left(\frac{a_t - |y_t|}{\tau}\right), \qquad
g_t = -s_t + (\pi - a_t)\,\frac{s_t(1-s_t)}{\tau},
$$
with a decaying learning rate $\delta_t = \delta_0/(t+1)^{0.5}$. The state transition is
$$
x_{t+1} = x_t + \delta_t\bigl(g_t + \varepsilon_t\bigr), \qquad \varepsilon_t \sim \mathcal{N}(0,\sigma_g^2).
$$

**Joint log-likelihood.** Because $x_{t+1} \mid x_t, y_t$ is normal, the full log-likelihood decomposes as
$$
\ell(\theta) =
\underbrace{\sum_{t=1}^{T}\log f_{\text{SDM}}(y_t \mid c,\kappa)}_{\text{response errors}}
\;+\;
\underbrace{\sum_{t=1}^{T-1}\log\phi\!\left(x_{t+1};\; x_t + \delta_t\, g_t,\; \delta_t\,\sigma_g\right)}_{\text{arc transitions}},
$$
with fitted parameter vector $\theta=(c,\kappa,\tau,\delta_0,\sigma_g)$. The exponent $\delta_{\text{pow}}=0.5$ is fixed. Below we verify identifiability via simulation-based parameter recovery and maximum-likelihood estimation with `optim`.

## Simulate data

Generate a single subject's worth of data from the noisy-gradient model with
known parameters. We use `run_arc_algorithm` which feeds SDM-generated response
errors `y` through the update function trial by trial.

```{r}
true_pars <- list(c = 6, kappa = 3, tau = 0.025, delta0 = 0.1, sigma_g = 2)

set.seed(4233)
sim <- run_arc_algorithm(
  arc_update_fun = algo5_noisy_grad(
    tau = true_pars$tau,
    delta0 = true_pars$delta0,
    delta_pow = 0.5,
    sigma_g = true_pars$sigma_g
  ),
  c = true_pars$c,
  kappa = true_pars$kappa,
  init_arc = pi/8,
  n_trials = 1000
)

par(mfrow = c(1, 3))
plot(sim$a, type = "l", xlab = "Trial", ylab = "Arc width")
hist(sim$a, xlab = "Arc width")
hist(sim$y, xlab = "Response error")
par(mfrow = c(1, 1))
```

## Log-likelihood

The joint log-likelihood has two additive components:

1. **SDM component** — the density of each response error $y_t$
2. **Arc-transition component** — a normal density on the logit scale for the
   arc update from trial $t$ to $t+1$

```{r}
algo5_loglik <- function(y, a, c, kappa, tau, delta0, delta_pow = 0.5, sigma_g) {
  T_ <- length(y)
  x <- logit(a / pi)

  # SDM log-likelihood for all y
  ll_y <- sum(dsdm(y, mu = 0, c = c, kappa = kappa, log = TRUE))

  # Arc-transition log-likelihood for t = 1..(T-1)
  # predicted x_{t+1} given (a_t, y_t)
  idx <- seq_len(T_ - 1)
  delta_t <- delta0 / (idx + 1)^delta_pow

  z <- (a[idx] - abs(y[idx])) / tau
  s <- 1 / (1 + exp(-z))
  grad <- -s + (pi - a[idx]) * s * (1 - s) / tau

  mu_x <- x[idx] + delta_t * grad
  sd_x <- delta_t * sigma_g

  ll_a <- sum(dnorm(x[idx + 1], mean = mu_x, sd = sd_x, log = TRUE))

  ll_y + ll_a
}
```

## Parameter transformations

All six parameters are strictly positive, so we estimate on the log scale and
exponentiate back.

```{r}
par_to_unconstrained <- function(pars) {
  log(unlist(pars[c("c", "kappa", "tau", "delta0", "sigma_g")]))
}

par_from_unconstrained <- function(theta) {
  vals <- exp(theta)
  list(c = vals[1], kappa = vals[2], tau = vals[3],
       delta0 = vals[4], sigma_g = vals[5])
}

neg_loglik <- function(theta, y, a) {
  pars <- par_from_unconstrained(theta)
  ll <- tryCatch(
    algo5_loglik(y, a, c = pars$c, kappa = pars$kappa, tau = pars$tau,
                 delta0 = pars$delta0, delta_pow = 0.5,
                 sigma_g = pars$sigma_g),
    error = \(e) -Inf,
    warning = \(w) -Inf
  )
  if (!is.finite(ll)) return(1e10)
  -ll
}
```

## Fit with `optim`

We initialise at perturbed values (not the truth) to verify that the optimiser
can recover the generating parameters.

```{r}
init_pars <- list(c = 3, kappa = 5, tau = 0.05, delta0 = 0.05, sigma_g = 1)
theta0 <- par_to_unconstrained(init_pars)

fit <- optim(
  par = theta0,
  fn = neg_loglik,
  y = sim$y,
  a = sim$a,
  method = "Nelder-Mead",
  control = list(maxit = 5000, reltol = 1e-10)
)

fit$convergence  # 0 = success
```

## Compare estimated vs true parameters

```{r}
estimated <- par_from_unconstrained(fit$par)

comparison <- tibble(
  parameter = names(true_pars),
  initial = unlist(init_pars),
  true_value = unlist(true_pars),
  estimated = unlist(estimated)
)
comparison
```

## Likelihood profile at the MLE

A quick sanity check: evaluate the likelihood over a grid around each MLE
estimate while holding the others fixed, to confirm we are at a peak.

```{r}
#| fig-width: 9
#| fig-height: 6
best_theta <- fit$par
par_names <- c("c", "kappa", "tau", "delta0", "sigma_g")

profiles <- purrr::map_dfr(seq_along(par_names), \(j) {
  grid <- seq(best_theta[j] - 1, best_theta[j] + 1, length.out = 80)
  ll_vals <- sapply(grid, \(v) {
    theta_tmp <- best_theta
    theta_tmp[j] <- v
    -neg_loglik(theta_tmp, y = sim$y, a = sim$a)
  })
  tibble(
    parameter = par_names[j],
    theta_unconstrained = grid,
    value = exp(grid),
    loglik = ll_vals
  )
})

ggplot(profiles, aes(x = value, y = loglik)) +
  geom_line() +
  geom_vline(data = comparison, aes(xintercept = estimated), linetype = "dashed", color = "blue") +
  geom_vline(data = comparison, aes(xintercept = true_value), linetype = "dotted", color = "red") +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "Parameter value", y = "Log-likelihood",
       caption = "Blue dashed = MLE, Red dotted = true value")
```

## Parameter recovery

Sample 10000 parameter sets uniformly, simulate data for each, fit via MLE, and
check whether the true values are recovered.

```{r}
#| label: param-recovery
n_recovery <- 10000
n_trials <- 1000
set.seed(2026)

recovery_true <- tibble(
  iter = seq_len(n_recovery),
  c = runif(n_recovery, 1, 10),
  kappa = runif(n_recovery, 1, 10),
  tau = runif(n_recovery, 0.001, 0.05),
  delta0 = runif(n_recovery, 0.01, 0.2),
  sigma_g = runif(n_recovery, 0, 10)
)

fit_one <- function(iter, c, kappa, tau, delta0, sigma_g) {
  sim <- run_arc_algorithm(
    arc_update_fun = algo5_noisy_grad(tau = tau, delta0 = delta0,
                                       delta_pow = 0.5, sigma_g = sigma_g),
    c = c, kappa = kappa, init_arc = pi / 2, n_trials = n_trials
  )

  theta0 <- par_to_unconstrained(list(
    c = 4, kappa = 4, tau = 0.02, delta0 = 0.08, sigma_g = 3
  ))

  fit <- tryCatch(
    optim(par = theta0, fn = neg_loglik, y = sim$y, a = sim$a,
          method = "Nelder-Mead",
          control = list(maxit = 5000, reltol = 1e-10)),
    error = \(e) list(par = rep(NA_real_, 5), convergence = 99)
  )

  est <- par_from_unconstrained(fit$par)
  tibble(
    iter = iter,
    c_est = est$c, kappa_est = est$kappa, tau_est = est$tau,
    delta0_est = est$delta0, sigma_g_est = est$sigma_g,
    converged = fit$convergence == 0
  )
}

recovery_est <- with_cache("output/algo5_recovery_est.rds", {
  purrr::pmap_dfr(recovery_true, fit_one)
})
```

Tag iterations where at least one estimate exceeds 4× the max true generating value:

```{r}
est_cols <- c("c_est", "kappa_est", "tau_est", "delta0_est", "sigma_g_est")
true_cols <- c("c", "kappa", "tau", "delta0", "sigma_g")
thresholds <- 4 * sapply(true_cols, \(col) max(recovery_true[[col]]))
outlier_mat <- sapply(seq_along(est_cols), \(j) {
  recovery_est[[est_cols[j]]] > thresholds[j]
})
recovery_est$bad <- rowSums(outlier_mat, na.rm = TRUE) > 0
```

Join true and estimated values and reshape for plotting:

```{r}
recovery_df <- recovery_true |>
  left_join(recovery_est, by = "iter") |>
  pivot_longer(
    cols = c(c, kappa, tau, delta0, sigma_g),
    names_to = "parameter",
    values_to = "true_value"
  ) |>
  mutate(
    estimated = case_when(
      parameter == "c" ~ c_est,
      parameter == "kappa" ~ kappa_est,
      parameter == "tau" ~ tau_est,
      parameter == "delta0" ~ delta0_est,
      parameter == "sigma_g" ~ sigma_g_est
    )
  ) |>
  select(iter, parameter, true_value, estimated, converged, bad)

cat("Converged:", sum(recovery_est$converged), "/", n_recovery,
    sprintf("(%.1f%%)", 100 * mean(recovery_est$converged)), "\n")
cat("Bad parameter estimates:", sum(recovery_est$bad), "/", n_recovery,
    sprintf("(%.1f%%)", 100 * mean(recovery_est$bad)), "\n")
```

Correlations between true and estimated values (converged runs only):

```{r}
recovery_df |>
  filter(converged, !bad) |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated),
    rmse = sqrt(mean((true_value - estimated)^2)),
    bias = mean(estimated - true_value),
    .groups = "drop"
  ) 
```

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery
#| fig-cap: "Parameter recovery for the noisy-gradient model. Each point is one simulated dataset. The dashed line is the identity."

plot_df <- recovery_df |> filter(converged, !bad)

cor_labels <- plot_df |>
  group_by(parameter) |>
  summarise(
    r = cor(true_value, estimated),
    x = -Inf, y = Inf,
    .groups = "drop"
  ) |>
  mutate(label = sprintf("r = %.2f", r))

plot_df |>
  ggplot(aes(x = true_value, y = estimated)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_point(aes(color = converged), alpha = 0.1, size = 1.5) +
  geom_text(data = cor_labels, aes(x = x, y = y, label = label),
            hjust = -0.1, vjust = 1.3, inherit.aes = FALSE, size = 4) +
  facet_wrap(~parameter, scales = "free") +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "red")) +
  labs(x = "True value", y = "Estimated value", color = "Converged")
```

Binned summary: for each parameter, true values are split into 20 equal-width
bins. Within each bin, we show the median estimate (line) and 80%/95% intervals
(ribbons).

```{r}
#| fig-width: 10
#| fig-height: 8
#| label: fig-recovery-binned
#| fig-cap: "Binned parameter recovery. Line = median estimate per bin; dark ribbon = 80% interval; light ribbon = 95% interval."

binned_df <- plot_df |>
  group_by(parameter) |>
  mutate(bin = cut(true_value, breaks = 20, labels = FALSE)) |>
  group_by(parameter, bin) |>
  summarise(
    true_mid = mean(true_value),
    median_est = median(estimated),
    q025 = quantile(estimated, 0.025),
    q10 = quantile(estimated, 0.10),
    q90 = quantile(estimated, 0.90),
    q975 = quantile(estimated, 0.975),
    .groups = "drop"
  )

ggplot(binned_df, aes(x = true_mid)) +
  geom_abline(linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = q10, ymax = q90), alpha = 0.3, fill = "steelblue") +
  geom_line(aes(y = median_est), color = "steelblue", linewidth = 0.8) +
  facet_wrap(~parameter, scales = "free") +
  labs(x = "True value", y = "Estimated value")
```


