---
title: "Honig et al.: Data Exploration"
author: "Ven Popov"
---

Honig et al. (2020) asked 12 participants to perform ~1200 trials of the arc-width task across four sessions, always with set size 4. Sessions 1 and 4 used uniformly distributed color stimuli; sessions 2 and 3 used normally distributed stimuli. This dataset provides a replication check and a within-subject reliability test because of the large trial count per participant.

This notebook is a preliminary exploration of the Honig dataset — observation counts and basic distributions. More detailed analyses using this data appear in the [sequential dependencies](sequential-dependencies.qmd) notebook.

::: {.callout-note}
## Data source
Preprocessed data: `output/honig_data.csv` (produced by `R/data-functions.R`).
:::

```{r warning=FALSE, message=FALSE}
#| label: setup
library(tidyverse)
```

This is what the data looks like:

```{r}
#| label: load-honig-data
honig_data <- read.csv("output/honig_data.csv")
```

Number of observations per subject and session:

```{r}
#| label: obs-by-session
honig_data |> 
  group_by(subject, session) |> 
  summarise(n = n(),
  b = 1) |> 
  pivot_wider(names_from = session, values_from = n)
```

note that sessions 1 and 4 use uniformly distributed stimuli, while sessions 2 and 3 use normally distributed stimuli.

Response error by subject

```{r}
#| label: resperr-by-subject
ggplot(honig_data, aes(resperr)) +
  geom_histogram() +
  facet_wrap(~subject) +
  theme_classic()
```
