---
title: "Explore reinforcement learning algorithms"
format: html
editor: visual
execute: 
  cache: true
---

```{r}
#| include: false
#| message: false
library(bmm)
library(ggplot2)
library(tidyr)
library(dplyr)

# the optimum alpha is the root of this equation
# it can be used with any distribution by providing the cdf and pdf functions
optimum_equation <- function(alpha, cdf, pdf) {
  (pi - alpha)*pdf(alpha) - cdf(alpha) + 0.5
}

# returns a data.frame with one row and two columns with the optimal alpha and phit
find_optimum_sdm <- function(c, kappa) {
  solution <- uniroot(
    optimum_equation, 
    cdf = \(x) psdm(x, c = c, kappa = kappa), 
    pdf = \(x) dsdm(x, c = c, kappa = kappa), 
    lower = 0, upper = pi-0.0001
    )
  data.frame(alpha = solution$root, phit = 2*psdm(solution$root, c = c, kappa = kappa)-1)
}
```

The analyses so far show that subject arc behavior is near optimal, if we assume perfect knowledge about the response error distribution. This doesn't mean that subjects have such knowledge. It might be possible to learn from the feedback on each trial and to adjust the arc width in a way that reaches optimal behavior.

For such an algorithm to work it would have to:

-   find the theoretical optimum or approximate it well enough to capture the empirical observation that optimal arc width increases with setsize and hit rate decreases with setsize

-   converge on the optimum fast enough (within the number of trials in the experiment; likely much earlier, so that the overall measure matches)

Below we explored some initial algorithms. We generate data from the SDM for a range of $c$ parameter values, and apply the learning algorithms to the trial-by-trial data generated from the model.

First we make a generic harness to run a trial-by-trial algorithm that updates the arc based on the function given in the first argument.

```{r}
run_arc_algorithm <- function(arc_update_fun, c = 4, kappa = 3, init_arc = pi/2, n_trials = 1e5) {
  y <- rsdm(n_trials, 0, c = c, kappa = kappa)

  a <- rep(0, n_trials)
  a[1] <- init_arc
  
  for (i in seq_len(n_trials-1)) {
    hit <- sign((abs(y[i]) <= a[i]) - 0.5)  # 1 if hit, -1 if miss
    a[i+1] <- arc_update_fun(a = a[i], hit = hit, y = y[i], i = i)
  }
  list(phit = mean(abs(y) <= a), y = y, a = a)  
}

# a convenience function to show how the algorithm performs on a single run
plot_algorithm_run <- function(out) {
  par(mfrow = c(1,2))
  plot(out$a, type = "l", xlab = "trial", ylab = "Arc half-width")
  plot(cummean(abs(out$y) <= out$a), type = "l", xlab = "trial", ylab = "Cummulative hit rate")
}
```

### Algorithm 1: Decrease/increase arc by delta depending on hit or not

If target is within arc, decrease arc by a proportion $\delta$ of the arc; otherwise increase by the same proportion

$$
a_{n+1} = a_n \pm \delta a_n = a_n (1 \pm \delta)
$$

```{r}
algo1 <- function(delta) \(a, hit, ...) a * (1 - hit*delta)

algo1(delta = 0.1) |> 
  run_arc_algorithm(n_trials = 1000) |>
  plot_algorithm_run()
```

Run for different c parameters

```{r}
cs <- seq(0, 10, 0.5)
out <- lapply(cs, \(c) run_arc_algorithm(algo1(delta = 0.1), c = c))

```

Distribution of arc widths for different $c$ values of the SDM parameter:

::: panel-tabset
```{r}
#| results: asis
for (i in seq_along(cs)) {
  cat("## c = ", cs[i], '\n\n')
  hist(out[[i]]$a, main = "", xlab = "Arcs chosen by algorithm", breaks = 60)
  cat("\n\n")
}
```
:::

Hit rate as a function of $c$ doesn't change.

```{r}
plot(cs, sapply(out, \(x) x$phit), ylim = c(0,1))
```

As a function of the step size $\delta$:

```{r}
deltas <- seq(0.01, 1, 0.01)
phit <- lapply(deltas, \(x) run_arc_algorithm(algo1(delta = x), c = 4)$phit)
plot(deltas, phit)
```

### Algorithm 2: Logarithmic decrease/increase arc by delta depending on hit or not

$$
a_{n+1} = a_n \left(1-\delta\right)^{\pm1}
$$

multiply by $(1-\delta)$ if hit, otherwise divide by it.

```{r}
algo2 <- function(delta) \(a, hit, ...) a * (1 - delta)^hit
```

One run of the algorithm:

```{r}
algo2(delta = 0.1) |> 
  run_arc_algorithm(n_trials = 1000) |>
  plot_algorithm_run()
```

```{r}
out <- lapply(cs, \(c) run_arc_algorithm(algo2(delta = 0.1), c = c))
plot(cs, sapply(out, \(x) x$phit), ylab = "Hit rate")

deltas <- seq(0.01, 0.5, 0.01)
phits <- lapply(deltas, \(x) run_arc_algorithm(algo2(delta = x), c = 4)$phit)
plot(deltas, phits)
```

### Algorithm 3: Change delta proportionally to the distance to the target

$$
a_{n+1} = a_n + \delta \, (|y_n|-a_n)
$$

```{r}
algo3 <- function(delta) \(a, hit, y, ...) a + delta * (abs(y) - a)

algo3(delta = 0.05) |> 
  run_arc_algorithm(n_trials = 1000) |> 
  plot_algorithm_run()
```

```{r}
phits <- lapply(cs, \(c) run_arc_algorithm(algo3(delta = 0.05), c = c)$phit)
plot(cs, phits)
```

With delta = 1 we get the case where you set the arc to be the same as the error:

```{r}
phits <- sapply(cs, \(c) run_arc_algorithm(algo3(delta = 1))$phit)
plot(cs, phits)
```

### Algorithm 4: Increase/decrease proportionally by arc remaining / arc used

$$
a_{n+1} = \cases{ 
   a_n - \delta \, a_n = a_n \left(1 - \delta \right) & if hit, \\
   a_n + \delta \left(\pi-a_n\right) = a_n \left(1 - \delta \right) + \delta \pi & otherwise
}
$$

```{r}
algo4 <- function(delta) \(a, y, ...) a * (1 - delta) + (abs(y) > a) * pi * delta

algo4(delta = 0.01) |> 
  run_arc_algorithm(n_trials = 1000) |> 
  plot_algorithm_run()
```

```{r}
cs1 <- seq(0, 10, 0.1)
deltas <- c(0.001, 0.05, 0.1, 0.2, 0.5)
pars <- expand.grid(c = cs1, delta = deltas)

out <- purrr::pmap(pars, \(c, delta) run_arc_algorithm(algo4(delta), c = c))
pars$phits <- sapply(out, \(x) x$phit)
pars$reward <- sapply(out, \(x) mean((pi - x$a)*(abs(x$y) <= x$a)))

sdm_optim <- lapply(cs1, \(c) find_optimum_sdm(c, kappa = 3))
sdm_optim <- do.call(rbind, sdm_optim)
sdm_optim$reward <- sdm_optim$phit * (pi - sdm_optim$alpha)

plot(cs1, sdm_optim$phit, type = "l", lty = "dashed", ylab = "Hit rate")
for (i in seq_along(deltas)) {
  lines(cs1, filter(pars, delta == deltas[i])$phit, col = i+1)
}
legend("topleft", legend = c("sdm optimal", deltas), col = seq_len(length(deltas)+1), lty = 1)

```

What this algorithm is doing is optimizing for the arc width a that satisfies:

$$
a = \pi (1 - F(a))
$$\
where F is the cdf of the sdm.\
\
This is similar to our objective but not exactly. That said, I decided to plot the achieved reward by the algorithm, rather than the hit rate, and compare it to the reward you get from the theoretical optimum:

```{r}
plot(cs1, sdm_optim$reward, type = "l", lty = "dashed", ylab = "Mean reward")
for (i in seq_along(deltas)) {
  lines(cs1, filter(pars, delta == deltas[i])$reward, col = i+1)
}
legend("topleft", legend = c("sdm optimal", deltas), col = seq_len(length(deltas)+1), lty = 1)
```

Despite the mismatching hit rate, it gets really close to the optimum reward, and only starts diverging more as c increases. This suggests that the gradient of the reward function might be gentle enough so that near the optimum point it can accomodate trade-offs between the arc and the hit rate with a huge decrement in the reward achieved.\
\
This will need more exploration but at least it seems like a plausible candidate (and/or with some modifications).
