---
title: "Explore reinforcement learning algorithms"
format: html
execute: 
  cache: true
---

```{r setup}
#| message: false
#| cache: false
# Load useful packages and custom functions
library(bmm)
library(ggplot2)
library(tidyr)
library(dplyr)
 
source("R/arc-models.R")
source("R/utils.R")
source("R/math.R")
```

The analyses so far [show that subject arc behavior is near optimal](optimal-arc-response.html), if we assume perfect knowledge about the response error distribution. This doesn't mean that subjects have such knowledge. It might be possible to learn from the feedback on each trial and to adjust the arc width in a way that reaches optimal behavior.

For such an algorithm to work it would have to:

-   find the theoretical optimum or approximate it well enough to capture the empirical observation that optimal arc width increases with setsize and hit rate decreases with setsize

-   converge on the optimum fast enough (within the number of trials in the experiment; likely much earlier, so that the overall measure matches)

Below we explored some initial algorithms. We generate data from the SDM for a range of $c$ parameter values, and apply the learning algorithms to the trial-by-trial data generated from the model. To see the code for the custom functions, please check the R folder on the github repository.

# Deterministic algorithms

In this section we explore algorithms that update the arc *deterministically* based on the observed arc response, distance to the target, and reward.

## Algorithm 1: Decrease/increase arc by delta depending on hit or not

If target is within arc, decrease arc by a proportion $\delta$ of the arc; otherwise increase by the same proportion

$$
a_{n+1} = a_n \pm \delta a_n = a_n (1 \pm \delta)
$$

```{r}
algo1(delta = 0.1) |> 
  run_arc_algorithm(n_trials = 1000) |>
  plot_algorithm_run()
```

Run for different c parameters

```{r}
cs <- seq(0, 10, 0.5)
out <- lapply(cs, \(c) run_arc_algorithm(algo1(delta = 0.1), c = c))
```

Distribution of arc widths for different $c$ values of the SDM parameter:

::: panel-tabset
```{r}
#| results: asis
par(mfrow = c(1,1))
for (i in seq_along(cs)) {
  cat("## c = ", cs[i], '\n\n')
  hist(out[[i]]$a, main = "", xlab = "Arcs chosen by algorithm", breaks = 60)
  cat("\n\n")
}
```
:::

Hit rate as a function of $c$ doesn't change.

```{r}
plot(cs, sapply(out, \(x) x$phit), ylim = c(0,1))
```

As a function of the step size $\delta$:

```{r}
deltas <- seq(0.01, 1, 0.01)
phit <- lapply(deltas, \(x) run_arc_algorithm(algo1(delta = x), c = 4)$phit)
plot(deltas, phit)
```

## Algorithm 2: Logarithmic decrease/increase arc by delta depending on hit or not

$$
a_{n+1} = a_n \left(1-\delta\right)^{\pm1}
$$

multiply by $(1-\delta)$ if hit, otherwise divide by it.

One run of the algorithm:

```{r}
algo2(delta = 0.1) |> 
  run_arc_algorithm(n_trials = 1000) |>
  plot_algorithm_run()
```

```{r}
out <- lapply(cs, \(c) run_arc_algorithm(algo2(delta = 0.1), c = c))
plot(cs, sapply(out, \(x) x$phit), ylab = "Hit rate")

deltas <- seq(0.01, 0.5, 0.01)
phits <- lapply(deltas, \(x) run_arc_algorithm(algo2(delta = x), c = 4)$phit)
plot(deltas, phits)
```

## Algorithm 3: Change delta proportionally to the distance to the target

$$
a_{n+1} = a_n + \delta \, (|y_n|-a_n)
$$

```{r}
algo3(delta = 0.05) |> 
  run_arc_algorithm(n_trials = 1000) |> 
  plot_algorithm_run()
```

```{r}
phits <- lapply(cs, \(c) run_arc_algorithm(algo3(delta = 0.05), c = c)$phit)
plot(cs, phits)
```

With delta = 1 we get the case where you set the arc to be the same as the error:

```{r}
phits <- sapply(cs, \(c) run_arc_algorithm(algo3(delta = 1))$phit)
plot(cs, phits)
```

## Algorithm 4: Increase/decrease proportionally by arc remaining / arc used

$$
a_{n+1} = \cases{ 
   a_n - \delta \, a_n = a_n \left(1 - \delta \right) & if hit, \\
   a_n + \delta \left(\pi-a_n\right) = a_n \left(1 - \delta \right) + \delta \pi & otherwise
}
$$

```{r}
algo4(delta = 0.01) |> 
  run_arc_algorithm(n_trials = 1000) |> 
  plot_algorithm_run()
```

```{r}
cs1 <- seq(0, 10, 0.1)
deltas <- c(0.001, 0.05, 0.1, 0.2, 0.5)
pars <- expand.grid(c = cs1, delta = deltas)

out <- purrr::pmap(pars, \(c, delta) run_arc_algorithm(algo4(delta), c = c))
pars$phits <- sapply(out, \(x) x$phit)
pars$reward <- sapply(out, \(x) mean((pi - x$a)*(abs(x$y) <= x$a)))

sdm_optim <- lapply(cs1, \(c) find_optimum_sdm(c, kappa = 3))
sdm_optim <- do.call(rbind, sdm_optim)
sdm_optim$reward <- sdm_optim$phit * (pi - sdm_optim$alpha)

plot(cs1, sdm_optim$phit, type = "l", lty = "dashed", ylab = "Hit rate")
for (i in seq_along(deltas)) {
  lines(cs1, filter(pars, delta == deltas[i])$phit, col = i+1)
}
legend("topleft", legend = c("sdm optimal", deltas), col = seq_len(length(deltas)+1), lty = 1)

```

What this algorithm is doing is optimizing for the arc width a that satisfies:

$$
a = \pi (1 - F(a))
$$

where F is the cdf of the sdm.

This is similar to our objective but not exactly. That said, I decided to plot the achieved reward by the algorithm, rather than the hit rate, and compare it to the reward you get from the theoretical optimum:

```{r}
plot(cs1, sdm_optim$reward, type = "l", lty = "dashed", ylab = "Mean reward")
for (i in seq_along(deltas)) {
  lines(cs1, filter(pars, delta == deltas[i])$reward, col = i+1)
}
legend("topleft", legend = c("sdm optimal", deltas), col = seq_len(length(deltas)+1), lty = 1)
```

Despite the mismatching hit rate, it gets really close to the optimum reward, and only starts diverging more as c increases. This suggests that the gradient of the reward function might be gentle enough so that near the optimum point it can accomodate trade-offs between the arc and the hit rate with a huge decrement in the reward achieved.

This will need more exploration but at least it seems like a plausible candidate (and/or with some modifications).

Here are more detailed individual runs:

```{r}
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(algorithm4 = algo4(delta = 0.01)),
  c = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi/2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
)
```

## Algorithm 5: Gradient ascent via smooth learner

Smooth approximation to hits to allow for gradient climbing (with $\tau$ being a tuning parameter):

$$
s = \frac{1}{1 + \exp\Bigl(-\frac{a-|y|}{\tau}\Bigr)}
$$

$s = 0.5$ if your arc response was right at the boundary; $s$ approaches 1 for hits further away from the boundary, and 0 for misses further away from the boundary.

Then you can get stochastic gradient ascent via the update rule:\
$$
a_{n+1} = a_n + \delta_i\Big[-s + (\pi-a)\frac{s(1-s)}{\tau}\Big]
$$

```{r}
algo5(tau = 0.025, delta0 = 0.1, delta_pow = 0.5) |>
  run_arc_algorithm(n_trials = 1000, init_arc = pi/2) |>
  plot_algorithm_run()
```

```{r}
taus <- c(0.01, 0.02, 0.05)
pars <- expand.grid(c = cs1, tau = taus)

out <- purrr::pmap(pars, \(c, tau) run_arc_algorithm(algo5(tau = tau, delta0 = 0.1, delta_pow = 0.5), c = c))
pars$phits <- sapply(out, \(x) x$phit)
pars$reward <- sapply(out, \(x) mean((pi - x$a)*(abs(x$y) <= x$a)))


plot(cs1, sdm_optim$phit, type = "l", lty = "dashed", ylab = "Hit rate")
for (i in seq_along(taus)) {
  lines(cs1, filter(pars, tau == taus[i])$phit, col = i+1)
}
legend("topleft", legend = c("sdm optimal", taus), col = seq_len(length(taus)+1), lty = 1)
```

More detailed individual runs:

```{r}
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(algorithm5 = algo5(tau = 0.025, delta0 = 0.1)),
  c = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = 1.0,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```

### Visualizing algorithm 5

The abstract form of the algorithm equation makes it difficult to see how it works. So let's visualize how the arc width will be updated depending on the distance between the arc and the response. Assume that the target is at a distance $\pi/2$ from the response, and vary the arc half width around the response. If the arc is less than $\pi/2$ then it's a miss (right side); otherwise it's a hit (left side)

```{r}
update_fun <- algo5(tau = 0.025, delta0 = 0.1, delta_pow = 0.5)

old_a <- seq(0, pi, 0.01)
new_a <- sapply(old_a, \(a) update_fun(a = a, y = pi/2, i = 1))
plot(old_a, new_a, type = "l", xlab = "Old arc", ylab = "New arc")
abline(v = pi/2, lty = 2)
abline(0, 1, col = "red")
```

That was unexpected. It is an interesting algorithm, which we can summarize as:

-   if it's a hit and the arc is much wider than the target, contract

-   if it's a miss and the arc is much smaller than the target, do nothing

-   if the arc boundary is close to the target, regardless of whether it's a miss or hit - expand.

The intuition is that if your arc is much smaller than the target, the fact that you miss can be due to chance - you know that even with a very precise memory some samples will be expected to deviate. When your arc is much bigger and you hit, you try to make it smaller. But if you are very close to the target, whether you hit or missed doesn't matter - a small change would have pushed you over from a hit to a miss or vice versa, and we know that there is sampling noise. So you increase the arc to be safer, proportionally to how close the call was.

I didn't design the algorithm this way, but it makes sense. For completeness, let's look at the same figure, but also varying where the target is. Above I used a target that is a quarter of the circle away.

```{r}
#| fig-width: 10
expand_grid(old_a = old_a, y = seq(0, pi, length.out = 9)) |>
  mutate(new_a = purrr::pmap_dbl(list(old_a, y), \(a, y) update_fun(a = a, y = y, i = 1)),
         y = round(y, 2)) |>
  ggplot(aes(x = old_a, y = new_a)) +
  geom_line(aes(color = factor(y))) +
  geom_abline(alpha = 0.5) +
  geom_vline(aes(xintercept = y), linetype = "dashed") +
  labs(x = "Old A", y = "New A", color = "Target") +
  facet_wrap(~y)

```

Right, so it's always the same principle, except that the amount of expansion when we are close to the target is biggest when the target was very close to the response, and smallest if it was on the opposite side of the circle. This makes sense - if the target is on the other side of the circle, and our arc is already big enough to be close to it, we don't expand much. But if the target is close to the origin, and our arc is really small, then expand a lot because "we are in danger". This fits the discussion we had about "the cost of errors" - when we have a small arc that is near optimal, changes to are a lot more costly, because an increment would lead to a lot more misses than the same increment if the optimal arc was much bigger.

# Stochastic algorithms

The two plausible algorithms that approximate the optimal reward behavior are algorithm 4 and algorithm 5. Both of them are deterministic - given a target and a produced arc on one trial, they will always predict the same arc for the next trial.

This makes fitting the models to data using likelihood-based methods impossible, because the likelihood will be 0 except at the exact parameter values that reproduce the arcs perfectly. In order to fit them to data, we have to decide on how to add stochasticity / noise to the predicted arcs.

To see why this is the case, let's formalize the general modeling framework explored so far in this notebook.

All models so far are examples of partially deterministic state-space models with the following structure:

-   observations:

$$
y_{1:T} \in [-\pi, \pi), \quad a_{1:T} \in [0, \pi)
$$

-   data model for the point response error (with parameter vector $\theta_y$):

$$
y_t \sim \text{SDM}(\theta_y)
$$

-   latent policy update (deterministic) with some update function $u$

$$
a_{t+1} = u(a_t, y_t, \theta_a)
$$

If we treat both $y$ and $a$ as generated by the model, the joint likelihood would be:

$$
p(y, a | \theta) = p(y | \theta_y) \cdot p(a | y, \theta_a)
$$

With a deterministic $u$ as in Algorithms 4 and 5, the second term on the right-hand side would be a product of Diract delta measures, which are 0 whenever their inputs are not 0:

$$
p(a | y, \theta_a) = \delta (a_1 - a_{init}) \prod_{t = 1}^{T-1} \delta (a_{t+1} - u(a_t, y_t, \theta_a))
$$

In order to make this into a proper likelihood, we need to add a stochastic component. The choice is not trivial, because it can alter the performance of the algorithm drastically, as we will see below; the choice will also impact how easy it is to estimate the models. We can distiguish between: intended (latent) arc, executed arc, where the noise enters, which of the two the agent remembers for updating, and which is used to calculate the reward.

All options must take one non-negotiable constraint into consideration - the reward shown to the agent / participants is based on the recorded response. There cannot be any noise that comes after the reward calculation but before the response recording in the system. Furthermore, we have to be careful about ensuring that all responses remain in $[0, \pi)$, either by clipping them, or by adding noise on an unconstrained transformation of the variable.

## Noise in execution with clipping

We let 

$$
a_{t+1} = u(a_t, u_t, \theta_a) + \epsilon_t
$$

where $\epsilon_t$ is normally distributed noise, and we restrict to $[0, \pi)$.

We fix the c parameter of the sdm to 8. For algorithm 4, adding randomly distributed noise around the intended arc response has a small-to-moderate effect on reducing the reward. But that effect seems to be reducing only the amplitude, and not the convergence properties of the algorithm. The plots below show the average reward, hit rate and arc width over trial averaged over 50 runs of the algorithm, with ribbons showing the 95% range of runs.

```{r}
clipped_noise <- function(..., sd, algo_fun) {
  min(pi, max(0, algo_fun(...) + rnorm(1, sd = sd)))
}

#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    algo4 = algo4(delta = 0.1), 
    algo4_0.05 = \(...) clipped_noise(..., sd = 0.05, algo_fun = algo4(delta = 0.1)),
    algo4_0.1 = \(...) clipped_noise(..., sd = 0.1, algo_fun = algo4(delta = 0.1)),
    algo4_0.2 = \(...) clipped_noise(..., sd = 0.2, algo_fun = algo4(delta = 0.1))
  ),
  cs = c(8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

That is not true for algorithm 5, whose performance drastically deteriorates with this type of noise. The noise added to the true response is likely interfering with the gradient estimation.

```{r}
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    algo5 = algo5(tau = 0.025, delta0 = 0.1),
    algo5_0.05 = \(...) clipped_noise(..., sd = 0.05, algo_fun = algo5(tau = 0.025, delta0 = 0.1)),
    algo5_0.1 = \(...) clipped_noise(..., sd = 0.1, algo_fun = algo5(tau = 0.025, delta0 = 0.1)),
    algo5_0.2 = \(...) clipped_noise(..., sd = 0.2, algo_fun = algo5(tau = 0.025, delta0 = 0.1))
  ),
  cs = c(8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

## Noise in execution (added to unconstrained variable)

To avoid the need to clip responses to be between 0 and pi, which will make estimation difficult, we can add the noise at a latent unconstrained version of the arc response. Since the arc varies from 0 to pi, we can use the logit function. Therefore:

$$
a_{t+1} = \sigma(\sigma^{-1}(u(a_t, u_t, \theta_a)/ \pi) + \epsilon_t)\pi
$$

where 

$$
\sigma(x) = 1 / (1 + \exp(-x))
$$

and

$$
\sigma^{-1}(p) = \log(p/(1-p))
$$

For algorithm 4, everything is stable as before, so we could use this for stable estimation:

```{r}
#| fig-asp: 0.5
logit_noise <- function(..., sd, algo_fun) {
  inv_logit(logit(algo_fun(...)/pi) + rnorm(1, sd = sd))*pi
}

ribbon_summary_runs(
  alg_list = list(
    algo4 = algo4(delta = 0.1), 
    algo4_noise_0.1 = \(...) logit_noise(..., sd = 0.1, algo_fun = algo4(delta = 0.1)),
    algo4_noise_0.2 = \(...) logit_noise(..., sd = 0.2, algo_fun = algo4(delta = 0.1)),
    algo4_noise_0.3 = \(...) logit_noise(..., sd = 0.3, algo_fun = algo4(delta = 0.1))
  ),
  cs = c(8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

Also for algorithm 5, there is no significant difference to clipping the noise on the native scale. The noise would have to be very tiny not to deteriorate performance.

```{r}
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    algo5 = algo4(delta = 0.1), 
    algo5_0.1 = \(...) logit_noise(..., sd = 0.1, algo_fun = algo5(tau = 0.025, delta0 = 0.1)),
    algo5_0.2 = \(...) logit_noise(..., sd = 0.2, algo_fun = algo5(tau = 0.025, delta0 = 0.1)),
    algo5_0.3 = \(...) logit_noise(..., sd = 0.3, algo_fun = algo5(tau = 0.025, delta0 = 0.1))
  ),
  cs = c(8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

## Noise gradient

Alternatively, we can add noise directly to the gradient. To make this clearer, here's a summary of the entire model.

### Deterministic ingredients

#### Soft hit indicator
Introduce a smooth proxy for $\mathbf 1\{|y_t|\le a_t\}$ with temperature $\tau>0$:
$$
z_t = \frac{a_t - |y_t|}{\tau},
\qquad
s_t = \sigma(z_t) = \frac{1}{1+\exp(-z_t)}.
$$

#### Gradient term
Define the gradient-like update term
$$
g(a_t,y_t;\tau)
= -s_t + (\pi-a_t)\,\frac{s_t(1-s_t)}{\tau}.
$$

#### Learning-rate schedule
Use a decaying step size
$$
\delta_t = \frac{\delta_0}{(t+1)^{\delta_{\text{pow}}}},
\qquad \delta_0>0,\ \delta_{\text{pow}}>0.
$$

### Noisy-gradient dynamics

A direct stochastic version of the update adds Gaussian noise to the gradient term. Because $a_t$ must remain in $(0,\pi)$, it is convenient to work on an unconstrained scale.

#### Reparameterization
Let
$$
x_t = \operatorname{logit}\!\left(\frac{a_t}{\pi}\right)\in\mathbb R,
\qquad
a_t = \pi\,\sigma(x_t).
$$

#### State transition
The stochastic transition is
$$
x_{t+1}
=
x_t + \delta_t\Big(g(a_t,y_t;\tau) + \varepsilon_t\Big),
\qquad
\varepsilon_t \sim \mathcal N(0,\sigma_g^2),
$$
with $a_t=\pi\sigma(x_t)$.

Equivalently, conditioning on $(x_t,y_t)$,
$$
x_{t+1}\mid x_t,y_t,\theta
\sim
\mathcal N\!\left(
x_t+\delta_t g(\pi\sigma(x_t),y_t;\tau),
\ (\delta_t\sigma_g)^2
\right),
$$
where the parameter vector is
$$
\theta = (\tau,\delta_0,\delta_{\text{pow}},\sigma_g).
$$

### Likelihood for fitting $(a,y)$

#### Arc component:

The likelihood for the arc component is:
$$
p(a_{2:T}\mid a_1,y_{1:T-1},\theta)
=
\prod_{t=1}^{T-1}
\mathcal N\!\left(
x_{t+1}\ \Big|\ x_t+\delta_t g(a_t,y_t;\tau),\ \delta_t\sigma_g
\right),
$$

where
$$
x_t = \operatorname{logit}(a_t/\pi),\quad a_t=\pi\sigma(x_t).
$$

#### Joint component (log likelihood)

If I also want to model the distribution of $y_t$ (with $\kappa$ fixed or estimated), then

Taking logs:
$$
\log p(y_{1:T},a_{2:T}\mid a_1,\theta)
=
\sum_{t=1}^T \log f_{\mathrm{SDM}}(y_t\mid 0, c, \kappa)
+
\sum_{t=1}^{T-1}
\log \phi\!\left(
x_{t+1};
x_t+\delta_t g(a_t,y_t;\tau),
\ \delta_t\sigma_g
\right).
$$

with $\phi$ being the log density of the normal distribution.

### Summary of fitted parameters

A minimal, well-posed parameterization is:
$$
\theta = (c, \kappa, \tau,\delta_0,\delta_{\text{pow}},\sigma_g),
$$

### Simulation

The `algo5_noisy_grad` function in `R/arc-models.R` implements this model. Unlike the logit-noise wrapper (which computes the deterministic update on the native scale and then perturbs on the logit scale), here the entire update---gradient plus noise---is applied on the unconstrained logit scale $x_t = \operatorname{logit}(a_t/\pi)$.

```{r}
#| fig-asp: 0.5
ribbon_summary_runs(
  alg_list = list(
    algo5_det = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0),
    noisy_grad_0.5 = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 0.5),
    noisy_grad_1 = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 1),
    noisy_grad_2 = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2),
    noisy_grad_5 = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 5)
  ),
  cs = c(8),
  n_trials = 1000,
  n_runs = 50,
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 1
) + ggplot2::theme(legend.position = "bottom")
```

Because the noise multiplies $\delta_t$ (which decays over trials), even substantial $\sigma_g$ values do not destroy convergence — the effective noise variance $(\delta_t \sigma_g)^2$ shrinks automatically with the learning rate. Compare this with the logit-noise execution variant, where fixed noise continually perturbs the arc and prevents convergence for algorithm 5.

Individual runs across several memory-strength values:

```{r}
#| fig-width: 9
#| fig-height: 15
spaghetti_single_runs(
  alg_list = list(noisy_grad = algo5_noisy_grad(tau = 0.025, delta0 = 0.1, sigma_g = 2)),
  cs = c(0, 2, 4, 8, 16),
  n_trials = 800,
  n_runs = 10,
  init_arc = pi / 2,
  plot_what = c("reward", "hit", "a"),
  add_oracle = TRUE,
  oracle_fun = \(c) find_optimum_sdm(c, kappa = 3),
  seed = 123
)
```


